---
title: "Shifting spaces: how to summarise multidimensional spaces occupancy? "
author: "Thomas Guillerme, Mark Puttick, Ariel Marcy, Vera Weisbecker"
bibliography: references.bib
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

<!-- When toggling to html, toggle snippet `compilation_html` below to eval = TRUE -->

<!-- To build the line numbered version (for submission) use the following yaml header snippet
output:
  pdf_document:
    fig_width: 8
    fig_height: 8
    keep_tex: true
  self_contained: false


output:
  html_document:
    fig_width: 8
    fig_height: 8
---
and edit the save .tex file to add the following

\usepackage{lineno} %in the package header

\linespread{2} %in the document header

\modulolinenumbers[1] % just after the \begin{document} tag
\linenumbers
 -->


```{r header, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
## Repeatability note:
## This whole paper is entirely reproducible and compiles as a single document. The data, figures
## and tables are all generated through the code snippets in this file. Note that the first time
## that this paper is compiled, it will generate the data which will take substantial time
## as the shiftingspace_supplementary.Rmd file completion takes ~10 minutes. Subsequent compilations
## will be much faster!

## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerme/moms"); library(moms)

## Setting the datapath
DATA_PATH <- "../data/processed/"

## Create data directory
if(!dir.exists(DATA_PATH)) {
    dir.create(path = DATA_PATH, showWarnings = FALSE)
}

## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues <- seq(15, 375, length = n + 1)
    grDevices::hcl(h <- hues, l = 65, c = 100)[1:n]
}

## Checking whether the data exists
if(!all(c("remove_05.Rda") %in% list.files(DATA_PATH))) {
    ## Run The supplementary material (simulation) ~ 30 minutes
    rmarkdown::render("shiftingspace_supplementary_simulation.Rmd", "html_document")    
}
if(!all(c("empirical_results.Rda") %in% list.files(DATA_PATH))) {
    ## Run The supplementary material ~ 15 minutes
    rmarkdown::render("shiftingspace_supplementary_empirical.Rmd", "html_document")
}
```


```{r compilation_html, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[2]] <- substitute(type <- ".png")
```

# Abstract

Multidimensional analysis of traits are an increasingly common tool-kit in ecology and evolution.
Such analysis are based on multidimensional datasets in which each dimension represents an original or transformed trait.
Thus, analyses summarise all observed and potential trait combinations in a multidimensional trait-space (e.g.
a morphospace or an ecospace).
Taxa will typically only occupy a subset of this space, so researchers apply one or more metrics to quantify the way in which organisms "inhabit" that trait-space.
By using these metrics, researchers have investigated how trait-space occupancy changes through time, in relation to other groups of organisms, and in response to global environmental changes, such as global warming events or mass extinctions.
<!-- AM: Perhaps add a “for example” somewhere as this doesn’t seem to be an exhaustive list (e.g. the trilobite paper we use looks more at change thru developmental events).  -->

Although a few space (well defined) occupancy metrics are used in ecology and evolution, we lack a general description of the majority of those metrics, especially in macroevolution.
In macroevolutionary studies these metrics are used as proxies for the concept of disparity.
Therefore the mathematical and biological meaning of such trait-space occupancy or disparity metrics can be vague.
This has consequences when interpreting changes in trait-space occupancy and linking them to biological phenomena.
For example, does a reduction in volume in trait-space correspond to episodes of extinction or changes in traits distribution? Does an overlap in trait-space mean overlapping niches?

Here we propose a user-friendly tool to Measure Occupancy in Multidimensional Spaces (moms) that allows visualisation of numerical changes of diverse metrics following space transformations (e.g. contraction, displacement, etc.).
It also allows the visualization of space changes through 2D projections.
This tool allows researchers to explore and visualise the properties of disparity metrics for their data under various parameters.
Furthermore, we propose a thorough review 
<!-- VW: This sounds too tacked on while it is a really important topic in its own right.  -->
of the behaviour of four common and two new metrics regarding various spatial properties (i.e. distribution, transformation, dimensionality).
As expected, we found that certain metrics are more appropriate than others in different disparity analyses.




# Introduction


Since the advent of modern biology,
<!-- AM: When is that exactly… can you pick a more helpful historical reference, like Linnaean Classification,  Darwinian Evolution or the Modern Synthesis?  -->
it has been acknowledged that groups of species and environments share specific, easily recognisable, correlated characteristics (e.g. guilds or biomes) of many kinds (e.g. phenotypic, physiological, phylogenetic, behavioural). 
Organisms or environments should therefore be studied as a set of traits rather
<!-- VW: Otherwise it sounds like people are going “We always knew this but oops! We’re only doing something about it now”. -->
 than some specific traits in isolation [e.g. @donohue2013; @hopkins2017].
increased computational power and statistical methods, as well as by the implementation of statistical methods in `R` packages [@oksanen2007vegan; @adams2013geomorph; @momocs; @blonder2018; @disprity] has facilitated a recent explosion of multiple traits studies approaches.


One popular way to analyse all these traits in concert is to use ordination (or dimensionality reduction) techniques [see @legendre2012 for a summary] to create a trait-space of the studied system.
<!-- AM: “Ordination is a popular way to analyse sets of traits in concert and create a trait-space of the study system [L&L 2012]. Ordination (also known as dimensionality reduction) has been used…”
Is it really “ordination techniques”? Or is ordination on its own sufficient? 
 -->
Such trait-space approach has been successfully used across many sub-disciplines in biology to either explore properties of the data or test hypothesis.
For example, in palaeobiology, such trait-spaces can be used to look how groups of species change in terms of shared characteristics through time [e.g. @wright2017];
in ecology, it can be used to compare two populations and see whether their overlap can be related to ecological processes such as competition [e.g. @jones2015].
However, different fields use a different set of terms and approaches for such trait-space approaches.
In macroevolution, for example, such trait-spaces are commonly referred to as "morphospaces" [with the whole procedure often being often called "disparity analysis"; e.g. @adams2013geomorph; @marcy2016; @wright2017], while in ecology, such trait-spaces can be called "functional spaces" [and "dissimilarity analysis"; e.g. @diaz2016; @mammola2019].
Nonetheless, they are the same mathematical objects: matrices with columns representing a trait (or a transformation thereof) and rows representing observations [taxon, field site, etc.; @disprity].
<!-- MP: It would be worthwhile to make a table showing how these same approaches are given different terminologies in each field. So rows with the approach and columns with ‘macroevolution’, ‘ecology’, etc., -->

Ecologists and evolutionary biologists are also using trait-spaces for similar reasons.
Across many sub-disciplines they will look at how their observations occupy the trait-space, often with respect to the same fundamental questions: are two groups overlapping in the trait-space?
Are some regions of the trait-space not occupied? Or how do specific factors influence the occupancy of the trait-space?
This is achieved in various ways such as using disparity metrics (or indices) in macroevolution [@wills2001; @hopkins2017; @disprity] or comparing hypervolumes in ecology [@donohue2013; @diaz2016; @blonder2018; @mammola2019].
Although these trait-space occupancy metrics are common in ecology and evolution, surprisingly little work has been published on their behaviour [but see @ciampaglio2001; @villéger2008; @mammola2019].
This can seem trivial but (unsurprisingly)
<!-- VW: Either you have a reference or you take out “unsurprisingly”. But this is your core idea, right? So I recommend phrasing it so it’s clear that this is the issue you have uncovered. -->
 different space occupancy metrics will capture different aspects of the trait-space.
For example, in many studies in ecology and evolution, authors have focused on measuring the volume of the trait-space [or a function thereof; e.g. ellipsoid volume @donohue2013; hypervolume @diaz2016; product of variance @wright2017; Procrustes variance @marcy2016] yet this only represents a single aspects of multidimensional occupancy that excludes other aspects that can be of interest [e.g. the density @geiger2008; the position @wills2001; @ciampaglio2001; or dissimilarity between groups @oksanen2007vegan; @close2015]
Therefore, this lack of diversity in multidimensional occupancy metrics may restrain the potential of multidimensional analysis [@villéger2008].
<!-- VW: Is there space for an easy-to-understand example that illustrates why volume of trait space might miss an important biological message when density is affected? -->

Here we propose a broad classification of space occupancy metrics as used across ecology [@villéger2008]  and evolution [@wills2001] and analyse their power to detect changes in multidimensional space occupancy in empirical and simulated data.
<!-- TG: Add something around here that our space transformation can be seen as "null" models. -->
<!-- VW: You are having the two issues (review/simulation) and moms) in different order compared to the abstract. The way you have it here makes more sense though! -->
We propose a review of each broad type of space occupancy metrics and a unified terminology aiming to foster easier communication between ecology and evolution.
Unsurprisingly, we found that there is no one-size-fits-all metric and that they are often dependent on the characteristics of the space and the question at hand.
However, because there can potentially be as many metrics as therea are trait-spaces and studies, it would be a daunting task to propose clear generalities to space occupancy metrics behavior.
Therefore, we propose a user friendly tool allowing researchers to design, experiment and visual their own space occupancy metric tailored for their specific project.

## Space occupancy metrics

For the purposes of this paper, we will call trait-spaces any matrix
<!-- AM: I’ve always thought of a morphospace as being the graphical representation of the underlying coordinate matrix. Add mention of graphical representation?  -->
 where rows are observations and columns are traits.
These traits can widely vary in number and types.
For example, traits could be discrete observations on an organism [e.g. presence of absence of a bone; @beck2014; @wright2017], continuous ones [e.g. leaf area; @diaz2016] or more sophisticated continuous ones [e.g. landmark position; @marcy2016].
Traits can also be relative observations such as community composition [e.g. @jones2015] or even some distance between observations [e.g. @close2015].
However, regardless of the methodology used to build a trait-space, three main broad occupancy metrics can be measured to investigate how observations are distributed in the trait-space: the volume which will approximate the amount of space occupied, the density which will approximate the distribution in space and the position which will approximate the location in space [Figure 1; @villéger2008].
Of course any combination of these three aspects is always possible.

```{r fig_metrics_types, echo = FALSE, fig.height = 3, fig.width = 9, results = 'hide', fig.cap = paste("Figure 1: different type of information captured by disparity metrics. A - Volume (e.g. sum of ranges); B - Density (e.g. average squared pairwise distances); C - Position (e.g. median distance from centroid).") }

set.seed(11)
## Plot space function (utility shortcut)
## The elements
elements <- 10

## Trait space
trait_space <- space.maker(elements, 2, distribution = rnorm)

## Graphical parameters
op <- par(mfrow = c(1,3), bty = "n")

## The volume
plot(trait_space, pch = defaults$pch, main = "A - Volume", xlab = "Trait 1", ylab = "Trait 2")
## The range lines
lines(x = range(trait_space[,1]), y = rep(mean(trait_space[,2]), 2), col = defaults$col1)
lines(x = rep(mean(trait_space[,1]), 2), y = range(trait_space[,2]), col = defaults$col1)

## The density
plot(trait_space, pch = defaults$pch, main = "B - Density", xlab = "Trait 1", ylab = "")
## Plotting the pairwise lines
pair.line <- function(line, trait_space, defaults) {
    lines(x = trait_space[line,1], y = trait_space[line,2], col = defaults$col1)
}
apply(combn(1:elements, 2), 2, pair.line, trait_space, defaults)
points(trait_space, pch = defaults$pch)

## The position
plot(trait_space, pch = defaults$pch, main = "C - Position", xlab = "Trait 1", ylab = "")
## Plotting the center of the space
points(0,0, pch = 13, cex = 2)
arrows(x0 = 0, y0 = 0, x1 = trait_space[,1], y1 = trait_space[,2], code = 0, col = defaults$col1)
points(trait_space, pch = defaults$pch)

par(op)
```

#### 1. The volume

Volume metrics measure the spread of a group
<!-- AM: Do we need to define that a “group” is not a subset of the set of traits, but is a group of observations grouped by another variable (e.g. species)?  -->
in the trait-space.
They can be interpreted as the amount of the trait-space that is occupied by observations.
Typically, larger values for such metrics indicate the presence of more extreme trait combinations in the sample.
For example, if group A has a bigger volume than group B, the observations in group A achieve more extreme trait combinations than in group B.
This type of metric is widely used in both ecology 
<!-- AM: What do ecologists use the hypervolume for? -->
[the hypervolume; @blonder2018] and in evolution [the sum or product of ranges or variances @wills2001; @ciampaglio2001 or the Procrustes variance @adams2013geomorph].
In general, such metrics are a variation of the hypercube volume.
<!-- VW: Should this go further up in this paragraph? 2nd sentence or so? -->
<!-- AM: Also, as it stands, I don’t know what this means. This is the first reference to hypercubes, which should probably be connected with the multi-dimensional set of traits (4+ dimensions) and the visualization challenges with that in the introduction to the introduction. Then another sentence in this paragraph can probably explain what a variation of the hypercube volume means   Also hypercube is similar to but distinct from the ecological method “hypervolume” which is mentioned in the previous sentence. -->


<!-- TODO:

’ -> '  
“ -> "
” -> "

-->




Although this type of metric
<!-- AM: I followed the daisy chain of “this type of metric” or “such metrics” back up to the first words of the previous paragraph: volume metrics. This seems circular! So I think I’m missing something  - perhaps see also comment re: hypercubes.  -->
 is a highly suitable indicator
<!-- AM: Citation needed? -->
  for comparing a group’s trait-space volume occupancy, 
<!-- AM: I think type of trait-space occupancy always has to be stated because "trait-space occupancy” in general is so central to this paper but distinctions between what is interesting is key to our message re: metric right for you -->
 it is limited to comparing the range of extreme trait-combinations between groups.
Typically, volume metrics do not take into account the spread of the observations within a group.
In other words, they can make it difficult to determine whether all the observations are on the edge of the volume or whether the volume is simply driven by small number of extreme observations.
<!-- AM: This seems quite important but is really buried in this paragraph. Potentially too much work… but have you considered a pro/con table of these categories of metrics. Could be helpful if our intended audience includes people who are not familiar with methods from other fields but want to use the moms paper to defend using a ecological method to ask a evolutionary question? -->
This “outlier bias” has been acknowledged in palaeoecology, for example for the product of ranges metric (the exact size of the hypercube) which is highly sensitive to outliers [@Butler2012]; or in ecology with the use of convex-hypervolumes.
<!-- AM: Citation -->
However, this “outlier bias” issue can be easily alleviated by using lower confidence intervals [e.g. by only at 95% of the data; @donohue2013 or by using the variance instead of the range; @ciampaglio2001]. 

<!-- AM: What does this mean? + Short paragraph about what questions are often/best tested with volume metrics? And/Or an example of what is missed if this is the only metric used? (same for density & position to reinforce main point of needing several) -->


#### 2. The density

Density metrics measure the distribution of a group in the trait-space.
They can be interpreted as an indication of the “packedness”
<!-- AM: Perhaps a sentence about how this is not the opposite of volume metric thing looking at extremes? -->
 of a group in trait-space.
For example, if group A is more dense than group B, observations in group A tend to be more similar between each other relatively than within group B.
Although density is an aspect of trait-space occupancy less routinely
<!-- AM: "Less routinely… still often” seems like an oxymoron -->
 measured, it is still often used in both evolution [e.g. the average pairwise distance; @geiger2008] and in ecology [e.g. the minimum spanning tree length; @oksanen2007vegan].

Despite their relatively rare application, density metrics could be used more often in concert with a volume metric to investigate finer hypotheses.
For example, if group A has a greater volume than group B but both have the same density, similar mechanisms could be driving both groups’ trait-space occupancy (e.g. a Brownian motion) but maybe group A is an older species and has had more time to evolve more extreme trait combinations.
<!-- VW: Good example but needs rewriting, I can do it but you said not to get too stuck into the wording. -->

#### 3. The position


Position metrics measure where a group lies in trait-space.
They can be interpreted as an indication of where a group lies in the trait-space either relative to the space itself or relative to another group.
For example, if group A has a different position than group B, observations in group A will have a different trait-combination than group B.

Position metrics may be harder to interpret in multidimensional spaces (i.e. beyond left/right and above/below) and are hence less often used.
However, when thinking about unidimensional 
<!-- AM: Can unidimensional data have either spread or volume?  -->
 data, this metric is obvious: for example, two groups A or B could have the same variance (i.e. “volume” or spread) with the same number of observations (i.e. the same density) but could have a totally different mean and thus be in different positions.
Although these metrics are rarer in macroevolution (to our knowledge), they have been used in ecology to compare the position of two groups relatively to each other [@mammola2019].




## One metric to rule them all?

In an ideal word, the choice of the metric should be based on the properties of changes in the trait-space of interest while taking into account the properties of the space (number of elements or/and dimensions and distribution of the data on each axis).
However, it is highly unlikely that such a one-size-fits-all metric exists; rather, it is important to assess the properties of a metric relative to the specific data used.
<!-- VW: Is this paragraph necessary? It seems to repeat the point. You could rename the title something like “benefits of considering multiple metrics” (but snazzier) and then a one liner “The use of multiple metrics to assess traitspace occupancy has the benefit of providing a more detailed characterization of occupancy changes. For example…” -->

If the question of interest is, say, to look at how trait-space occupancy changes in response to mass extinction, using a single metric to describe one aspect of occupancy (usually the volume) can miss an important part of the picture.
For example, in @halliday2015, the authors demonstrate a change in mammalian trait-space volume as a result of the Cretaceous-Palaeogene extinction
<!-- VW: Hmm. Any chance we could use a different paper? Or phrase it differently in terms of “the data also seem to contain interesting information on trait space position…” -->
 (66 million years ago - Mya - Fig. 5 in @halliday2015) yet this seems to be also accompanied by a shift in the trait-space position due to the extinction of basal mammals (Fig. 8 in @halliday2015).
Such examples can be found throughout the literature in both evolution and ecology on different aspects of trait-space occupancy [e.g. @close2015; @diaz2016].
Figure 2 illustrates this problem on a 2D trait space with eight hypothetical observation where the three aspects (volume, density and position) can be decoupled and change in concert
<!-- AM: Another oxymoron -->
 or independently.(in empirical data, the relations can be more complex to discern).
Yet, this illustrates how using only one specific space occupancy metric (e.g. the volume) will fail to capture changes in the trait-spaces C, D, and I.
<!-- VW: Hmm. Does there even need to be this figure? I mean, isn’t it obvious? Or is it obvious but no one has thought of it? -->


<!-- In Slater et al 2017: variation in mean squared pairwise distance (density) with two peaks (25 Mya and 5 Mya) corresponding to a less densily packed trait-space (Fig. 5) which might be not optimal to explain the shift in body mass (position) and the increase in variance (volume) for the one trait analysed (body length).
 -->


<!-- Figure 2: Illustration how three different volume (Vol. - Product of ranges), density (Den. - Average nearest neighbour distance) and position (Pos. - Average displacement) metrics capture different occupancy aspects in this simplified trait-space (Numbers indicate hypothetical multiples of change? ). -->
<!-- VW: The caption probably needs to be a little more detailed. -->

```{r fig_metric_captures, echo = FALSE, fig.height = 4, fig.width = 8, results = 'hide', fig.cap = paste("Figure 2: Illustration how three different volume (Vol. - Product of ranges), density (Den. - Average nearest neighbour distance) and position (Pos. - Average displacement) metrics capture different occupancy aspects in this simplified trait-space.") }

## Making the volume/density/position space
vdp_space <- moms::vdp.make()

## Measuring disparity
vdp_disp <- moms::vdp.dispRity(vdp_space,
                         volume = c(prod, ranges),
                         density = c(mean, neighbours),
                         position = c(mean, displacements))

## Checking the table of volume changes
silent <- moms::vdp.check.table(vdp_disp, vdp_space)

## Plotting the results
moms::vdp.plot(vdp_space, disparity = vdp_disp,
        plot.names = c("A - Base (no changes)",
                       "B - Change in volume",
                       "C - Change in density",
                       "D - Change in position",
                       "E - Change in vol. & den.",
                       "F - Change in vol. & pos.",
                       "I - Change in den. & pos.",
                       "J - All changes"))
```

To allow a more confident assessment of whether trait space changes occur, and also to refine the characterization of trait space change, we suggest using multiple metrics to describe trait-space occupancy the same way that multiple statistics are required to describe distributions.
For example, during a radiation following a mass extinction event, one could both predict an increase of the volume of the trait-space [e.g. an increase in morphological diversity in birds; @crouch2019] but also a shift if position in the trait-space [(]e.g. a shift from non-avian to avian dinosaur traits; @crouch2019].
Similarly, in ecology, differences in trait-space density with constant volume could indicate selection pressures on the studied traits (i.e. due to the “narrowness” of trait combination for example in desert plants).
<!-- VW: This needs to be clearer but it’s great to have the example there. -->

In this paper, we analyse the effect of changes in simulated and empirical trait-spaces on seven space occupancy metrics to explicitly measure trait-space density and position. As expected, we found no one-size-fits-all metric but provide guidance on which metric could be used in which specific scenario. We also propose a new user-friendly software with an intuitive graphical user interface (GUI) to look at the effect of changes in morphospace
<!-- AM: Trait-space? Or specifically morphospace? -->
 occupancy on user defined disparity metrics: [`moms`]().
<!-- VW: This needs to be more aggressive but I can help with that when you want deeper text changes. -->








<!--  -->
<!--  -->
<!--  -->
<!-- METHODS SECTION -->
<!--  -->
<!--  -->
<!--  -->








# Methods

We tested how seven different space occupancy metrics 1) relate to each other, 2) are affected by modifications of traits space and 3) affect group comparisons in empirical data.
To do so we performed the following steps (explained in more details below):

1. We simulated 13 different spaces with different set of parameters (dimensions, distributions, etc);
2. We transformed these spaces by removing 50% of the observations following four different scenarios: randomly, by limit, by displacement and by density;
3. We measured occupancy on the resulted transformed spaces using seven different space occupancy metrics;
4. We applied the same space occupancy metrics to six empirical datasets from the literature (covering a range of disciplines and a range of dataset properties).

| Test | Methods section | Results
|------|-----------------|------------------
| How do the metrics compare to each other? | Metrics comparisons       | Figure 4
| How are they affected by space shifting?  | Space shifting            | Table 5
| How is that reflected in the real world?  | Empirical examples        | Table 6

Table 1: summary of the tests in this paper.

## Generating spaces

We generated trait spaces using the following combinations of size, distributions, variance and correlation:

| space name | size  | distribution(s)                | dimensions variance | correlation |
|------------|-------|--------------------------------|---------------------|-------------|
| Uniform3   |200*3  | Uniform (min = -0.5, max = 0.5)| Equal            | none        |
| Uniform15  |200*15 | Uniform                        | Equal            | none        |
| Uniform50  |200*50 | Uniform                        | Equal            | none        |
| Uniform150 |200*150| Uniform                        | Equal            | none        |
| Uniform50c |200*50 | Uniform                        | Equal            | Random (between 0.1 and 0.9) |
| Normal3    |200*3  | Normal  (mean = 0, sd = 1)     | Equal            | none        |
| Normal15   |200*15 | Normal                         | Equal            | none        |
| Normal50   |200*50 | Normal                         | Equal            | none        |
| Normal150  |200*150| Normal                         | Equal            | none        |
| Normal50c  |200*50 | Normal                         | Equal            | Random (between 0.1 and 0.9) |
| Random     |200*50 | Normal, Uniform, Lognormal (meanlog = 0, sdlog = 1)| Equal            | none        |
| PCA-like   |200*50 | Normal                         | Multiplicative           | none        |
| PCO-like   |200*50 | Normal                         | Additive              | none        |
Table 2: different simulated space distributions.

The different in space sizes (200 $\times$ 3, 200 $\times$ 15, 200 $\times$ 50 or 200 $\times$ 150) reflects the range of trait-space dimensions in literature: "low-dimension" spaces ($<15$) are common in ecology [@mammola2019] whereas high dimension spaces ($>100$) are common in macroevolution [@hopkins2017].
We used a different range of distributions (uniform, normal or random) to test the effect of observation distributions on the metrics.
We used different levels of variance for each dimensions in the spaces by making the variance on each dimension either equal (i.e. $\sigma_{D1} \simeq \sigma_{D2} \simeq \sigma_{Di}$) or decreasing (i.e. $\sigma_{D1} < \sigma_{D2} < \sigma_{Di}$) with the decreasing factor being either multiplicative (using the cumulative product of the inverse of the number of dimensions: $\prod_i^d(1/d)$) or additive (using the cumulative sum of the inverse of the number of dimensions: $\sum_i^d(1/d)$).
Both multiplicative and cumulative reductions of variance are used to illustrate the properties of ordinations where the variance decreases per dimensions [in a lognormal way in Principal components analysis - PCA; e.g. @marcy2016; healy2019; and in a normal way in Multidimensional Scaling - MDS, PCO or PCoA; e.g. @close2015; @wright2017].
Finally we added a correlation parameter to take into account the potential correlation between different traits.
We repeated the simulation of each trait-space 20 times (resulting in 260 trait-spaces).

## Spatial occupancy metrics

We then measured height @@@ different metrics on the resulting transformed spaces:

| Name | Definition | Captures | Source | Notes  |
|------|------------|----------|--------|--------|
| Average distance from centroid | $\frac{\sqrt{\sum_{i}^{n}{({k}_{n}-Centroid_{k})^2}}}{d}$ | Volume | @laliberté2010 | This is equivalent to the functional dispersion (FDis) in @laliberté2010 (but not weighted by observation's abundance) |
| Sum of variances | $\sum_{i}^{d}{\sigma^{2}{k_i}}$ | Volume | @wills2001 | This is a really common metric used in palaeobiology [@ciampaglio2001] |
| Sum of ranges | $\sum_{i}^{d}{\|\text{max}(d_{i})-\text{min}(d_{i})\|}$ | Volume | @wills2001 | This metric is more sensitive to outliers than the sum of variances |
| Ellipsoid volume | $\frac{\pi^{d/2}}{\Gamma(\frac{d}{2}+1)}\displaystyle\prod_{i}^{d} (\lambda_{i}^{0.5})$ | Volume | @donohue2013 | This metric is less sensitive to outliers than the convex hull hypervolume [@diaz2016; @blonder2018] |
| Minimum spanning tree average distance | $\frac{\sum(\text{branch length})}{n}$ | Density | @oksanen2007vegan | This metric is similar to the unscaled functional evenness [@villéger2008] |
| Minimum spanning tree distances evenness | $\frac{\sum\text{min}\left(\frac{\text{branch length}}{\sum\text{branch length}}\right)-\frac{1}{n-1}}{1-\frac{1}{n-1}}$ | Density | @villéger2008 | This metric is the functional evenness without weighted abundance [FEve; @villéger2008] |
| Average nearest neighbour distance | $min\left(\sqrt{\sum_{i}^{n}{({q}_{i}-p_{i})^2}}\right)\times \frac{1}{n}$ | Density | @foote1990 | This metric measures the density of pairs of observations in the trait-space (c.f. the Minimum spanning tree average distance above) |
| Average displacement | $\frac{\sqrt{\sum_{i}^{n}{({k}_{n})^2}}}{\sqrt{\sum_{i}^{n}{({k}_{n}-Centroid_{k})^2}}}$ | Position | This paper | This is the ratio between the observations' position from their centroid and the centre of the trait-space. A value of 1 indicates that the observations' centroid is the centre of the trait-space |
Table 3: List of metric with *n* being the number of observations, *d* the total number of dimensions, *k* any specific row in the matrix, *Centroid* being their mean and $\sigma^{2}$ their variance. $\Gamma$ is the Gamma distribution and $\lambda_{i}$ the eigen value of each dimension and ${q}_{i}$ and $p_{i}$ are any pairs of coordinates.


```{r metrics_list, echo = FALSE}
## Loading the list of metrics
source("list.of.metrics.R")
```

We selected these height @@@ space-occupancy metrics to illustrate how they capture different aspects of trait-space occupancy.
We do not believe that they are better than many of the other metrics used in ecology and evolution.
The supplementary results contain the same analysis as described below performed using more metrics.
Additionally, the [`moms`](@@@) shiny App allows to explore the effect of many more metrics as well as customising metrics by combining them together or entering users functions.

<!--  -->
<!--  -->
<!--  -->
<!-- COMPARING METRICS -->
<!--  -->
<!--  -->
<!--  -->


## Metrics comparisons

We compared the space occupancy correlation between each pair of metrics to check if they captured similar signal [@villéger2008; @laliberté2010].
First, we used the metrics on the full 13 trait-spaces described above.
We then scaled the results and measured the pairwise Pearson correlation to test whether metrics were capturing a similar signal (high positive correlation), a different signal (correlation close to 0) or an opposite signal (high negative correlations).
We performed all the tests using the `psych` package [@psych].

<!--  -->
<!--  -->
<!--  -->
<!-- EFFECT OF DIMENSIONALITY -->
<!--  -->
<!--  -->
<!--  -->
<!--
## Effect of dimensionality

> VW: I understand teh below but only just and it is hard to read. I can help simplify.

Furthermore, high dimensional spaces ($>> 3$) are subject to the "curse of multidimensionality" [@cursedimensionality].
This is link to the fact that the more dimensions, the more sparse the data becomes.
In other words the probability of two points A and B overlapping in *n* dimensions is the probability of point A having the same value as point point B on dimension 1, 2, 3, etc...
Thus this probability decreases as a product of the number of dimensions.
This "curse" can make the interpretation of high dimensional data counter-intuitive.

For example, the product of range is expected to decrease with the number of dimensions, regardless of the number of elements on each dimensions.
This is counter-intuitive because:
1) first, the range of values on a dimension is bounded to the number of elements on this dimensions (i.e. increasing the number of observations on one dimension can only increase the range, never decrease it);
2) second, products are a classical example of the "curse" where multiplying a higher number of dimensions with a non-null probability of having a range of zero will irremediably result in 0 (i.e. an infinite number of positive values by 0 is still 0).

-->

```{r fig_curse, eval=FALSE, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 3: Effect of the curse of dimensionality on different scaled disparity metrics. Metrics with containing products (ellipsoid volume, product of ranges or variances) are classically affected by the \"curse\" as they decrease more or less rapidly with an increase in dimensions. Note especially how the Product of Range actually increases and then decreases with the number of dimensions.")}
set.seed(42)
## Make a 50 dimensions space
space <- space.maker(300, 50, rnorm, scree = rev(cumsum(rep(1/50, 50))))

## Function wrappers
disparity <- function(metric, dimensions, space) {
    fast.disparity(group = 1:nrow(space), space = space, metric = metric, rare.dim = dimensions)
}
lapply.fun <- function(dimensions, metrics_list, space) {
    unlist(lapply(metrics_list, disparity, dimensions = dimensions, space = space))
}

## Curse results
curse_results <- sapply(2:50, lapply.fun, metrics_list = metrics_list, space = space)
curse_results <- apply(curse_results, 1, function(x) return(x/max(x)))

## Empty plot
hue_colors <- default.palette(ncol(curse_results))
par(bty = "n")
plot(NULL, xlim = c(2,50), ylim = c(0,1),
xlab = "Dimensions (decreasing variance)",
ylab = "Scaled disparity")
for(metric in 1:ncol(curse_results)) {
    lines(curse_results[, metric], col = hue_colors[metric])
}
legend("bottomleft", metric_names, lty = 1, col = hue_colors)
```
<!--  -->
<!--  -->
<!--  -->
<!-- SHIFTING SPACES -->
<!--  -->
<!--  -->
<!--  -->

## Shifting spaces {#shifting-spaces}

To measure how the metrics responded to shifts in trait-spaces, we reduced the spaces by removing 50% of elements each time using the following algorithms:

* **Randomly:** by randomly removing 50% of elements (Fig. 3A).

* **Limit:** by removing all elements with a distance from the centre (mean point) of the space lower or greater than a radius $\rho$ (where \$rho$ is selected such that 50% elements are selected) (Fig. 3B). This generated two limit removals: *maximum* and *minimum* respectively in orange and blue in Fig. 3B.

* **Density:** by removing any pairs of point with a distance $D$ from each other where (where $D$ is selected such that 50% elements are selected - Fig. 3C). This generated two density removals: *high* and *low* respectively in orange and blue in Fig. 3C.

* **Position:** by removing points similarly as for **Limit** but using the distance from the furthest point from the centre (max point - Fig. 3D). This generated two position removals: *positive* and *negative* respectively in orange and blue in Fig. 3D.


The algorithm to select $\rho$ or $D$ is described in more detail in in the [Appendix](#Appendix_algorithm_reduce).
We used mirror results for the limit, displacement and density reduction (i.e. the opposite data was also used).
For example, when increasing the **density** by removing the 50% most distant elements, we also run a transformation by removing 50% of the *closest* elements.


```{r fig_reduce_space, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 3: different type of space reduction. Each panel displays two groups (orange and blue) of 50% of the data points each. Each groups are generated using the following algorithm: A - randomly; B - by limit (maximum and minimum limit in respectively orange and blue); C - by density (high and low in respectively orange and blue); and D - by position (positive and negative in respectively orange and blue).")}
set.seed(42)

## Change colors
defaults$col1 <- "blue"
defaults$col2 <- "orange"

op <- par(mfrow = c(2,2), bty = "n")
## The elements
elements <- 300
## The amount to remove
remove <- 0.5

## Trait space
trait_space <- space.maker(elements, 2, distribution = rnorm)

## The randomly removed space
random_rm <- reduce.space(trait_space, type = "random", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, random_rm, main = "A - Random removal", defaults)

## The limit removal
limit_rm <- reduce.space(trait_space, type = "limit", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, limit_rm, main = "B - Limit", defaults)

## The density removal
density_rm <- reduce.space(trait_space, type = "density", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, density_rm, main = "C - Density", defaults)

## The displacement removal
displace_rm <- reduce.space(trait_space, type = "displacement", remove = 0.5)
## Moving the space to the upper right corner (positive)
plot.space(trait_space, displace_rm, main = "D - Position", defaults)


## List of removals (for plotting the tables later)
typical_removal <- list(space = trait_space,
                        remove = list("random" = random_rm,
                                      "limit(min)" = limit_rm,
                                      "limit(max)" = !limit_rm,
                                      "density(high)" = density_rm,
                                      "density(low)" = !density_rm,
                                      "displace(positive)" = displace_rm,
                                      "displace(negative)" = !displace_rm))
```

To measure the effect of space reduction, distribution and dimensionality on the metric, we first scaled the metric to be relative to the non reduced space for each dimension distributions or number of dimensions.
To do so, we subtracted the observed occupancy with no space reduction (i.e. base occupancy) to all the occupancy measurements of the reduced spaces and then divided it by the resulting maximum observed occupancy.
This way, our occupancy metrics where scaled between -1 and 1 with a value of 0 indicating no effect of the space reduction and $>0$ and $<0$ respectively indicating an increase or decrease in the occupancy metric value.
We then measured the probability of overlap of the between the non-random removals (limit (max/min), displacement (positive/negative) and density (high/low)) and the random removals using the Bhattacharrya Coefficient [i.e. the probability of overlap between two distributions; @bhattacharyya1943; @guillerme2016].

### Measuring the effect of space and dimensionality on shifting spaces

We also measured the potential effect the distribution of the space or the number of dimensions in the space could have on the metric in general.
In fact, the difference of distributions can have some drastic effect on the way space-occupancy is measured.
For example, in a normally distributed space, a decrease in density can often lead to an increase in volume.
This is not necessarily true in log-normal spaces or in uniform spaces for certain metrics (e.g. the convex hull of a lognormal space will be less sensitive to changes in density).

Furthermore, the differences in distributions can have some drastic effects in high dimensional spaces ($> 10$) which are subject to the "curse of multidimensionality" [@cursedimensionality].
This is linked to the fact that the more dimensions, the more sparse the data becomes.
In other words the probability of two points A and B overlapping in *n* dimensions is the product of the probability of the two points overlapping on each dimensions ($\prod_{i}^{d} P(A = B)_{Di}$)
Thus this probability decreases as a product of the number of dimensions.
This "curse" can make the interpretation of high dimensional data counter-intuitive.

We thus measured the effect of space distribution and dimensionality using an ANOVA ($occupancy \sim distribution$ and $occupancy \sim dimensions$) by using 1) all spaces with 50 dimensions and 2) the uniform and normal spaces with equal variance and no correlation wit 3, 15, 50, 100 and 150 dimensions (Table 2) for testing respectively the effect of 1) distribution and 2) dimensions.
The results of the ANOVAs (_p_-values) are reported in Table 3 (see supplementary material for the full ANOVA results).

<!--  -->
<!--  -->
<!--  -->
<!-- EMPIRICAL EXAMPLES -->
<!--  -->
<!--  -->
<!--  -->

## Empirical examples

Finally, we analysed the effect of the different space-occupancy metrics on six different empirical studies covering a broad representative view of fields using trait-spaces analysis (palaeobiology, macroevolution, evo-devo, ecology, etc.)
For each of these six studies we generated a trait-spaces from the data published along with the paper.
We then divided the trait-space into two roughly equal sized groups based on the nature of the paper and applied the space-occupancy metrics to each pair of groups to test whether different metrics captured different differences between groups.

The procedures to generate the data and the groups varies from one study to the other but is detailed and fully reproducible in the supplementary materials.

study | field | taxonomic Group | traits (data) | trait-space | size | groups | type of question |
------|-------|-----------------|---------------|-------------|------|--------|------------------|
@beck2014 | palaeontology | Mammalia | discrete morphological phylogenetic data | Ordination of a distance matrix (PCO) | 106*105 | 52 crown vs. 54 stem | Are crown mammals more disparate than stem mammals?|
@wright2017 | palaeontology | Crinoidea | discrete morphological phylogenetic data | Ordination of a distance matrix (PCO) | 42*41 | 16 before vs. 23 after | Is there a difference in disparity before and after the Ordovician mass extinction?|
@marcy2016 | geometrics morphometrics | Rodentia | skull 2D landmarks coordinates  | Ordination of a Procrustes Superimposition (PCA) | 454*134 | 225 *Megascapheus* vs. 229 *Thomomys* | Are two genus of Gophers (*Megascapheus*, *Thomomys*) morphological distinct? |
@hopkins2016 | geometrics morphometrics | Trilobita | 3D landmarks | Ordination of a Procrustes Superimposition (PCA) | 46*46 | 10 juveniles vs. 36 adults | Are juveniles trilobites a subset of adult ones? |
@jones2015 | Ecology | Plantae | Communities species compositions | Ordination of a Jaccard distance matrix (PCO) | 48*47 | 24 aspens vs. 24 grasslands | Is there a difference in species composition between aspens and grasslands? |
@healy2019 | Ecology | Animalia | Life history traits | Ordination (PCA) | 285*6 | 83 ecthotherms vs. 202 endotherms | Do endotherms have more diversified life history strategies than ectotherms? |

Table 4: details of the six empirical trait-spaces.

For each empirical trait-space we bootstrapped each group 500 times [@disprity] and applied the height @@@ space-occupancy metric to each pairs of groups.
Note that when space-occupancy could not be calculated (e.g. due to the curse of multidimensionaly), we collapsed the metric value to 0 (i.e. no space-occupancy was captured).
We then compared the means of each groups using the Bhattacharrya Coefficient [i.e. the probability of overlap between two distributions; @bhattacharyya1943; @guillerme2016].
The are displayed in Table 6.

<!--  -->
<!--  -->
<!--  -->
<!-- RESULTS SECTION -->
<!--  -->
<!--  -->
<!--  -->








# Results

```{r loading_results, echo = FALSE}
## Loading the results
remove_05 <- load.results("remove_05")
```

```{r running_tests, echo = FALSE}
## Anova function
anova.fun <- function(data) {return(aov(glm(disparity ~ factor, data = data)))}

## Running the tests
all_test <- test.simulation(remove_05, test = anova.fun, scale = TRUE)
all_dim_test <- test.simulation(remove_05, test = anova.fun, scale = TRUE,
                      factors = c("uniform3", "uniform15", "uniform50", "uniform100", "uniform150",
                                  "normal3", "normal15", "normal50", "normal100", "normal150"))
space_test <- test.simulation(remove_05, test = anova.fun, scale = TRUE,
                              factors = c("uniform50", "uniform50c", "normal50", "normal50c",
                                          "random50", "pca_like", "pco_like"))
```

## Metrics comparisons 

```{r fig_metric_correlation, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 4: pairwise correlation between the scaled metrics. Numbers on the upper right corner are the Pearson correlations. The red line are linear regressions (with the confidence intervals in grey).")}

## Adding proper names
results_pairwise <- remove_05
names(results_pairwise[[1]]) <- metric_names

## Plotting the pairwise results
pairwise.plot(results_pairwise, scale = TRUE, type = "base", plot = "cor")
```

All the metrics selected where either positively correlated (correlation of 0.99 for the Average distance from centroid and Sum of variance or 0.97 for the Average nearest neighbour distance and Minimum spanning tree average length; Fig. 4) or somewhat correlated (ranging from 0.66 for the Sum of Variances and the Ellispsoid Volume to -0.09 between the Average displacement and the Average distance from centroid; Fig. 4).
Note that all metrics but the Ellipsoid Volume were normally (or nearly normally) distributed (Fig. 4).
This is due to the Ellipsoid Volume being more sensitive to the curse of multidimensionality (many values close to 0).
More comparisons between metrics are available in the supplementary materials.

## Space shifting

```{r fable_results, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## The metrics names (shortened vector)
name <- metric_names

## Making a list of parameters for each mini plot
plot.param <- list(scaler = 3,
                   bg.col = "black",
                   col = c("grey", "orange", "blue"),
                   quantiles = c(95, 50),
                   cent.tend = median,
                   pch = 19,
                   metric.max = length(metrics_list),
                   cex = 2)

## Looping through each mini plot for every metric
for(metric in 1:length(metrics_list)) {
    generate.fable.plot(data = remove_05, metric = metric, what = "limits", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = remove_05, metric = metric, what = "densit", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = remove_05, metric = metric, what = "displa", plot.param = plot.param, overlap = TRUE)
}

```

```{r, echo = FALSE, eval = FALSE}
## Function for printing the table in the R console
print.fable <- function(n_metrics, byrow, ncol, test) {
  ## Make the plot.id table 
  ids <- matrix(1:(n_metrics*ncol), ncol = ncol, byrow = byrow)

  for(one_metric in 1:n_metrics) {
    ## Print all the rows one by one
    text <- c(paste0("`r name[", one_metric, "]`"), paste0("`r plot.id(", ids[one_metric, ], ")`"))
    ## Add some tests?
    if(test) {
      text <- c(text, paste0("`r s.test(", one_metric, ", \"s\")`"), paste0("`r s.test(", one_metric, ", \"r\")`"))
    }
    ## Print the line
    cat(paste(text, collapse = " | "))
    cat("|\n")
  }
}
## Getting the fable to copy paste under the table header below.
print.fable(length(metrics_list), byrow = TRUE, ncol = 3, test = TRUE)
```

Metric      | Volume change  | Density change | Position change | Distribution effect | Dimensions effect |
:-----------|----------------|-----------------|----------------|---------------------|-------------------|
`r name[1]` | `r plot.id(1)` | `r plot.id(2)` | `r plot.id(3)` | `r s.test(1, "s")` | `r s.test(1, "r")`|
`r name[2]` | `r plot.id(4)` | `r plot.id(5)` | `r plot.id(6)` | `r s.test(2, "s")` | `r s.test(2, "r")`|
`r name[3]` | `r plot.id(7)` | `r plot.id(8)` | `r plot.id(9)` | `r s.test(3, "s")` | `r s.test(3, "r")`|
`r name[4]` | `r plot.id(10)` | `r plot.id(11)` | `r plot.id(12)` | `r s.test(4, "s")` | `r s.test(4, "r")`|
`r name[5]` | `r plot.id(13)` | `r plot.id(14)` | `r plot.id(15)` | `r s.test(5, "s")` | `r s.test(5, "r")`|
`r name[6]` | `r plot.id(16)` | `r plot.id(17)` | `r plot.id(18)` | `r s.test(6, "s")` | `r s.test(6, "r")`|
`r name[7]` | `r plot.id(19)` | `r plot.id(20)` | `r plot.id(21)` | `r s.test(7, "s")` | `r s.test(7, "r")`|
`r name[8]` | `r plot.id(22)` | `r plot.id(23)` | `r plot.id(24)` | `r s.test(8, "s")` | `r s.test(8, "r")`|


Table 5: Results of the effect of space reduction, space dimensions distributions and dimensions number of the different space-occupancy metrics. The dots represent the median space-occupancy values, the solid and dashed line respectively the 50% and 95% confidence intervals. Results in grey are the random 50% reduction. Results in blue are the minimum limit, low density and negative shift for respectively the volume, density and position changes. Results in orange are the maximum limit, high density and positive shift for respectively the volume, density and position changes. The displayed value in the plots are the probability of overlap (Bhattacharrya Coefficient) between the blue and orange reductions and the random ones. _p_-values for distribution effect and dimensions effect represents respectively the effect of the ANOVAs space-occupancy ~ distributions and space-occupancy ~ dimensions (Signif. codes:  0 '\*\*\*' 0.001 '\*\*' 0.01 '\*' 0.05 '.' 0.1 ' ' 1).

As expected from the sketch in Figure 2, some different metrics capture different aspects of trait-space occupancy.
However, it is worth noticing that it can be hard to predict the behaviour of each metric when 50% of the observations are removed.
In fact we observe a clear decrease in median metric in less than a third of the space reductions (10/36).

In terms of change in volume (increase or decrease), only the Average distance from centroid and the Sum of variances seem to capture a clear change in both directions.
Note however that the increase in volume does not correspond to an *actual* increase in volume in the trait-space (i.e. the volume in from the orange observations in Fig. 3B is equivalent to the one in Fig. 3A).
In terms of change in density (increase or decrease), only the Minimum spanning tree average distance and the Average nearest neighbour distance seem to capture a clear change in both directions.
In terms of change in position (increase or decrease), only the Average displacements metric seems to capture a change a clear change in direction (albeit not in both directions).
This is not so surprising since the notion of positions becomes more and more complex to grab as dimensionality increases (i.e. beyond left/right and up/down).

## Empirical example

```{r test_empirical, echo = FALSE}
## Loading the results
empirical_results <- load.results("empirical_results")

## Testing the differences for each distributions
bhatt.coeff.safe <- function(x, y, tol = 1e-16, ...) {
  if(all(x < tol) | all(y < tol)) {
    bhatt.coeff(1, 0)
  } else {
    bhatt.coeff(x, y, ...)
  }
} 
disparity_test <- lapply(empirical_results, lapply, test.dispRity, test = bhatt.coeff.safe)
```

```{r fable_results_empirical, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Plotting parameters
plot.param <- list(cex = 2,
                   col = c("#F7B27E", "#BFE4E3"),
                   border = c("#F65205", "#3E9CBA"),
                   na.cex = 3,
                   scaler = 3)

## Looping through each mini plot for every metric
data_names <- c("Beck and Lee 2014", "Wright 2017", "Marcy et al. 2016",
                "Hopkins and Pearson 2016", "Jones et al. 2015", "Healy et al. 2019")

for(dataset in 1:length(data_names)){
    for(metric in 1:length(metrics_list)){
        ## Plotting the results
        generate.fable.empirical(data = empirical_results[[dataset]][[metric]],
                                 test = disparity_test[[dataset]][[metric]],
                                 precision = 1e-5, plot.param, dataset = dataset)
    }
}
```

```{r change_plotid_chain, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[3]] <- substitute(chain <- "fable_results_empirical")
```

```{r, echo = FALSE, eval = FALSE}
print.fable(length(metrics_list), byrow = FALSE, ncol = 6, test = FALSE)
```

Metric     | Beck and Lee 2014  | Wright 2017 | Marcy et al. 2016 | Hopkins and Pearson 2016 | Jones et al. 2015 | Healy et al. 2019  |
:-----------|----------------|-----------------|------------------|----------------|------------------|----------------|
`r name[1]` | `r plot.id(1)` | `r plot.id(9)` | `r plot.id(17)` | `r plot.id(25)` | `r plot.id(33)` | `r plot.id(41)`|
`r name[2]` | `r plot.id(2)` | `r plot.id(10)` | `r plot.id(18)` | `r plot.id(26)` | `r plot.id(34)` | `r plot.id(42)`|
`r name[3]` | `r plot.id(3)` | `r plot.id(11)` | `r plot.id(19)` | `r plot.id(27)` | `r plot.id(35)` | `r plot.id(43)`|
`r name[4]` | `r plot.id(4)` | `r plot.id(12)` | `r plot.id(20)` | `r plot.id(28)` | `r plot.id(36)` | `r plot.id(44)`|
`r name[5]` | `r plot.id(5)` | `r plot.id(13)` | `r plot.id(21)` | `r plot.id(29)` | `r plot.id(37)` | `r plot.id(45)`|
`r name[6]` | `r plot.id(6)` | `r plot.id(14)` | `r plot.id(22)` | `r plot.id(30)` | `r plot.id(38)` | `r plot.id(46)`|
`r name[7]` | `r plot.id(7)` | `r plot.id(15)` | `r plot.id(23)` | `r plot.id(31)` | `r plot.id(39)` | `r plot.id(47)`|
`r name[8]` | `r plot.id(8)` | `r plot.id(16)` | `r plot.id(24)` | `r plot.id(32)` | `r plot.id(40)` | `r plot.id(48)`|

Table 6: Comparisons of different pairs of groups in different empirical trait-spaces. NAs are used for cases where space-occupancy could not be measured due to the curse of multidimensionality. The displayed values are the probability of overlap between both groups (Bhattacharrya Coefficient).

Similarly as for the simulated results, the empirical ones indicate that there is no perfect one-size-fit all metric.
For all height @@@ metrics (expect the Ellispoid volume) we see either one group or the other having a bigger mean than the other.
When looking at the probability of overlap, they are also clearly affected by the different metrics.


<!--  -->
<!--  -->
<!--  -->
<!-- DISCUSSION SECTION -->
<!--  -->
<!--  -->
<!--  -->








# Discussion

As expected we see a range of differences between the metrics themselves (Fig. 4), their response to space shifting (Tab. 5) and their effect on capturing differences in empirical datasets (Tab. 6).
This should not come as a surprise because a space-occupancy metric is bound to capture different summary of the trait-spaces depending on both the metric and the trait space properties.

#### Metrics comparisons

The correlation between metrics ranges from a near 1:1 correlation (0.99 for the mean distance from centroid and the sum of variances) to a somewhat negative correlation (-0.23 for the sum of ranges and the spanning tree average length).
Furthermore, all metrics seems to behave following a normal distribution but for the ellipsoid volume which demonstrates a classic "curse of multidimensionality" with values increasing logarithmically first and then quickly decreasing to 0 [Fig. 4; @cursedimensionality].
In other words, the ellipsoid volume is really sensitive to changes in trait-space only up until a modest number of dimensions.
The fact that we have such a range of correlations for such conserved distributions (but for the ellipsoid volume) hints that each metric can capture different summaries of trait-space occupancy ranging from obvious differences (for metrics not strongly correlated) to subtle ones (for metrics strongly correlated).
Comparisons between more metrics are available in the Supplementary Material 4@@@.

#### Space shifting

Most metrics capture a normal "null" space shifts with no changes in space occupancy on average (i.e. randomly removing 50% of the data - in grey in Table 5).
This is a desirable behaviour for space occupancy metrics since most studies using trait-space occupancy look at potential processes resulting in the observed space occupancy patterns [e.g. competition @brusatte2008; convergence @marcy2016; life history traits @healy2019; etc.].
In this context, random changes in the trait-space occupancy should not be picked up as a pattern to avoid false positive errors.
However, the average nearest neighbour distance and the sum of ranges have a respectively positive and negative "null" median meaning that any decrease of elements will always lead to respectively and increase and a decrease of these metrics even when the process is random (Table 5).
This is not especially a bad thing for both metrics (i.e. they can still be used to analysis changes in space occupancy) but should be kept in mind as a property of these metrics.

Regarding the changes in volume of the trait-spaces, the sum of variances and the average distance from centroid are really good descriptors with both maximum and minimum limit changes leading to a clear increase or decrease of the metrics values.
However, as illustrated in the 2D examples in Fig. 3B, only the minimum limit changes (orange) should actually lead to a decrease of volume of the trait-space; the volume of the maximum limit shifted trait-spaces should have the same volume as the non-modified trait-space but with a hole in its centre.
This could be due to the fact that these metrics are merely approximating changes in volume with the volume itself (e.g. the ellispsoid volume) being more sensitive to the curse of multidimensionality [@cursedimensionality].

Regarding changes in density of the trait-spaces, the average nearest neigbhour distance and the minimum spanning tree average distance consistently pick a change in density with more precision for low density trait-spaces (in blue in Table 5).
However these metrics seem to be also associated to a "null" increase (see above).
In general we can observe some degree of correlation between the changes in density and the changes in volume for most metric picking either signal (i.e. the average nearest neighbour distance picks both changes in density and in volume).
This could be due the generation of normally distributed spaces where a change in density often leads to a change in volume (i.e. to increase density in a normal distribution, one can simply reduce the its tails).

Regarding the changes in position of the trait-space, most metrics but the average displacements metric seems to not be able to distinguish between a random change in trait-space (grey) and a positive or a negative displacement of the trait-space (blue and orange - Table 5).
Furthermore the average displacements metric does not distinguish between and positive or a negative displacement of the trait-space: in both cases, the metric increases.
This might be due to the inherent complexity of "position" in a trait space that increases with the number of dimensions.
In fact, it is complex to intuitively describe position beyond a 2D (left/right) or 3D (up/down) space.

#### Empirical examples

Again, as expected, different metrics highlight different differences for different trait-spaces.
Although most differences are fairly consistent within each dataset (red group mean superior/inferior to blue group mean), this can be more or less pronounced (ranging from no to nearly full overlap - BC $\in(0;0.995)$) and sometimes reversed.
For example, in @wright2017 dataset where the blue group has a greater occupancy metric median value but for the average displacements metric showing a higher median displacement in the red group (Table 6).
These differences in group's space occupancy depending on the metrics are also more pronounced in the empirical datasets where the elements per group is highly different [@hopkins2016; @healy2019].
This also highlights the influence of different groups size on top of the differences in trait spaces and metrics properties.
Finally it is worth noting that the ellispoid volume only picks up accurate differences between groups in the @healy2019 dataset which is yet another illustration of the curse of multidimensionality: the @healy2019 dataset has only 6 dimensions compared to all the other ones having more than 41 dimensions [and the @marcy2016 dataset having 136 dimensions for which the ellipsoid volume can't be calculated; Table 6; @cursedimensionality].

## Caveats

Simulating trait spaces is complex.
For example the trait spaces are simulated by treating each observation as independent.
This is not often the case in biology since observations might be spatially [@jones2015] or phylogenetically auto-correlated [e.g. @beck2014].
However we can note that the we get expected results from the empirical data (where the observations are more realistically auto-correlated).
Furthermore, the algorithm used to reduce the multi-dimensional spaces might not always accurately reflect how trait-spaces are changed in nature.
They might indeed favour the response of specific metrics, in particular for the changes in density that only modifies the density between pairs of points rather than changing the global density.
This is specific algorithmic choice was made in order to not confound changes in density along with changes in volume (i.e. a general change in density can be obtain by uniformly increasing/decreasing the volume).

## Suggestions

We insist that there is no metric better than another one and that researchers should use the most appropriate metrics based on their question and knowledge of both the metric and the trait-space properties.
However, following the present study we suggest the several points that could be applied to potentially improve future multidimensional analysis:

First, we suggest using multiple metrics to tackle different aspects of the trait-space. 
This is suggestion is in the same logical thinking line that the mean can not be sufficient to describe a distribution (the variance and the type of distribution might be good additional indicators).
Although using multiple metrics is not uncommon in macroevolutionary studies [e.g. @halliday2015] or in ecology [@mammola2019], they often do not cover contrasted aspects of the trait-space (and are often different proxies of the volume of a trait-space).

Second, we suggest selecting a metric (or a series of metrics) that best help answering the biological question a hand.
In fact, if one studies an adaptive radiation in a group of organisms, it is worth thinking what would be the expected null model: would the group's volume increase (radiation in all directions), would it increase in density (niche specialisation) or would it shift in position (radiation into a new set of niches)?

Third, we suggest to not name metrics as the biological aspect they are potentially describing (e.g. disparity or functional dispersion) but rather what they are measuring (e.g. sum of dimensions variance or the average distance from centroids).
We believe this point will allow both a clearer understanding of what is measured (c.f. what is approximated) and a easier understanding between ecology and evolution for which the metrics can be the same but named differently and used for approximating different things.
For example, the average squared pairwise distance [`geiger::dtt`; @geiger2008] and the Procrustes Variance [`geomorph::disparity`; @adams2013geomorph] often describe the same aspect of space occupancy (see Supplementary material 4@@@).

# Conclusion

STILL DRAFT:
Multidimensional analysis is the future in biology.
It is important however to know what patterns we are describing from multidimensional data.
There are non one-size-fits-all space-occupancy metric.
Check the properties of your space and your metric (using [`moms`](@@@)!).
Use multiple metrics.


# Acknowledgements

Thanks to Natalie Jones for helping finding/sorting the ecological data.
Money from the ARC DP170103227, FT180100634

# References



# Supplementary material

## Algorithm for selecting the parameters to reduce the space ($radius$, $displacement$, $density$) {#Appendix_algorithm_reduce}

We used a recursive algorithm for selecting the parameter that removes $P$ = 50% elements.

1. Select a random reduction parameters $R$.
2. Remove elements from the space using $R$ resulting in $P'$ removed elements.
If the remaining number of elements is to the required proportion $P$ ; exit the algorithm;
Else go to 3.
3. Get the different between the proportion of removed elements $P'$ and $P$.
If the difference is positive set the increment parameter $R$ to $R = 1.1 \times R$, then go to 2.
Else set $R = 0.9 \times R$, then go to 2.

The algorithm is implemented in the `optimise.parameter` function in [reduce.space_fun.R](@@@).
