---
title: "Shifting spaces: which disparity or dissimilarity measurement best summarise occupancy in multidimensional spaces?"
author: "Thomas Guillerme, Mark N. Puttick, Ariel E. Marcy, Vera Weisbecker"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_width: 12
    fig_height: 6
---

# Supplementary material 5: effect of sampling on space occupancy measurements
 
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerm/moms"); library(moms)

## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues = seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[1:n]
}
```

In this section, we look at the effect of sampling on the 25 selected space occupancy measurements.
We modified the protocol presented in the main text as follows:

1.  We simulated 13 different spaces with different sets of parameters;
2.  We transformed these spaces by removing 50% of the observations following four different scenarios corresponding to different empirical scenarios: randomly, by limit (e.g. expansion or reduction of niches), by density (e.g. different degrees of competition within a guild) and by position (e.g. ecological niche shift).
2-bis. We reduced the sampling of the 50% selected elements by selecting 100% of them (50% of the total), 80% (40% of the total), 60% (30% of the total), 40% (20% of the total) and 20% (10% of the total).
3.  We measured occupancy on the resulting transformed spaces using `r metric.names()` different space occupancy measures;
4.  We applied the same space occupancy measures to six empirical datasets with the similar re-sampling strategy as in point 2-bis.

# Data

## Simulations

```{r start_simulation, echo = TRUE, eval = FALSE}
## Set the overall number of replicates
n_replicates <- 20

## Set the overall number of elements
#elements <- function() sample(20:200) #possibility to use a variable number of elements
elements <- function() 200

## Simulation seed
set.seed(42)
```

```{r generate_spaces_simulation, echo = TRUE, warning = FALSE, results = 'hide', eval = FALSE}
## List of distributions
distributions_list <- list(
    "unifor" = list(distribution = runif, arguments = list(list("min" = -0.5,
                                                                "max" = 0.5))),
    "normal" = list(distribution = rnorm),
    "random" = list(distribution = "random"),
    "unicor" = list(distribution = runif, arguments = list(list("min" = -0.5,
                                                                "max" = 0.5)),
                    cor.matrix = "random"),
    "norcor" = list(distribution = rnorm, cor.matrix = "random"),
    "pcalik" = list(distribution = rnorm, scree = "lognormal"),
    "pcolik" = list(distribution = rnorm, scree = "normal")
    )

## Generate all 13 spaces
all_spaces <- list(uniform3 = space.simulation(distributions_list$unifor,
                                               dimensions = 3,
                                               elements = elements(),
                                               replicates = n_replicates),
                   uniform15 = space.simulation(distributions_list$unifor,
                                               dimensions = 15,
                                               elements = elements(),
                                               replicates = n_replicates),
                   uniform50 = space.simulation(distributions_list$unifor,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates),
                   uniform100 = space.simulation(distributions_list$unifor,
                                               dimensions = 100,
                                               elements = elements(),
                                               replicates = n_replicates),
                   uniform150 = space.simulation(distributions_list$unifor,
                                               dimensions = 150,
                                               elements = elements(),
                                               replicates = n_replicates),
                   uniformc50 = space.simulation(distributions_list$unicor,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates),
                   normal3 = space.simulation(distributions_list$normal,
                                               dimensions = 3,
                                               elements = elements(),
                                               replicates = n_replicates),
                   normal15 = space.simulation(distributions_list$normal,
                                               dimensions = 15,
                                               elements = elements(),
                                               replicates = n_replicates),
                   normal50 = space.simulation(distributions_list$normal,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates),
                   uniform100 = space.simulation(distributions_list$unifor,
                                               dimensions = 100,
                                               elements = elements(),
                                               replicates = n_replicates),
                   normal150 = space.simulation(distributions_list$normal,
                                               dimensions = 150,
                                               elements = elements(),
                                               replicates = n_replicates),
                   normalc50 = space.simulation(distributions_list$norcor,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates),
                   random50 = space.simulation(distributions_list$random,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates),
                   pca_like = space.simulation(distributions_list$pcalik,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates),
                   pco_like = space.simulation(distributions_list$pcolik,
                                               dimensions = 50,
                                               elements = elements(),
                                               replicates = n_replicates))

#TG: Warnings are due to inexact correlations
```


```{r shift_spaces_simulation, echo = TRUE, warning = FALSE, results = 'hide', eval = FALSE}
## Shifting each space
shift_groups <- lapply(all_spaces, lapply, shift.group.simulation, remove = 0.5)
```

```{r get_disparity_simulation, echo = TRUE, warning = FALSE, results = 'hide', eval = FALSE}
## Measuring disparity
remove_05 <- mapply(metrics.simulation, all_spaces, shift_groups,
                  MoreArgs = list(metrics = metrics_list, rare.dim = NULL,
                                  verbose = TRUE),
                  SIMPLIFY = FALSE)

## Saving the results
save.results(remove_05)
```

## Empirical data


```{r}
## Results list
save.results(empirical_results)
```

# Results
