---
title: "Shifting spaces: how to summarise multidimensional spaces occupancy?"
author: "Thomas Guillerme, Mark Puttick, Ariel Marcy, Vera Weisbecker"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Supplementary material 1: simulations

```{r, echo = FALSE}
## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerm/moms"); library(moms)

## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues = seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[1:n]
}
```

# Simulation data

This section contains all the results and calculations (and plots them).
The whole analysis takes approximatively 30 minutes to run (single core 2.2 GHz).

## Metrics analysed

```{r metrics_list, echo = FALSE}
## All metrics
metrics_list <- list(
    "av.pairwise" = function(matrix) return(mean(pairwise.dist(matrix)^2)),
    "Procrustes" = function(matrix) return(sum(matrix^2)/nrow(matrix)),
    "Volume" = ellipse.volume,
    "PoR" = c(prod, ranges),
    "SoR" = c(sum, ranges),
    "PoV" = c(prod, variances),
    "SoV" = c(sum, variances),
    "PoQ" = c(prod, quantiles),
    "SoQ" = c(sum, quantiles),
    "av.displac" = c(mean, displacements),
    "av.neighbo" = c(mean, neighbours)
)

metric_names <- c("Ellipsoid volume", "Average squared distance", "Procrustes variance",
                  "Product of Range", "Sum of Ranges", "Product of Variances", "Sum of Variances",
                  "Product of Quantiles", "Sum of Quantiles", "Mean displacement",
                  "Mean nearest neighbours distance")
```

The metrics analysed are the `r paste(metric_names, collapse = ", ")`.

## Generating the data

This section runs and summarises the simulations.
The simulations are run silently in this markdown file (and are not printed on the compiled version of this document).
The code is available within this R markdown file (un-compiled).

```{r toggle_simulations, echo = FALSE}
## This arguments allows to toggle the data generation on (TRUE) or off (FALSE)
generate_data <- TRUE

## Set the overall number of replicates
n_replicates <- 20

## Simulation seed
set.seed(42)
```

```{r run_metrics, echo = FALSE, warning = FALSE, results = 'hide'}
#, results = 'hide'}
if(generate_data) {
    ## Bundle simulations 
    remove_05 <- simulation.spaces(remove = 0.5, replicates = n_replicates,
                                   metrics_list = metrics_list)
    # remove_08 <- simulation.spaces(remove = 0.8, replicates = n_replicates,
    #                                metrics_list = metrics_list)
    # remove_02 <- simulation.spaces(remove = 0.2, replicates = n_replicates,
    #                                metrics_list = metrics_list)

    ## Save simulations
    save.results(remove_05)
    # save.results(remove_08)
    # save.results(remove_02)
} else {
    ## Load results
    remove_05 <- load.results("remove_05")
    # remove_08 <- load.results("remove_08")
    # remove_02 <- load.results("remove_02")
}

## Summarise the data
remove_05 <- lapply(remove_05, summarise.metrics)
# remove_08 <- lapply(remove_08, summarise.metrics)
# remove_02 <- lapply(remove_02, summarise.metrics)
```

```{r run_rarefaction_dimensions, echo = FALSE, warning = FALSE, results = 'hide'}
#, results = 'hide'}
if(generate_data) {
    ## Bundle simulations
    rarefy_dim_05 <- simulation.rarefaction(remove = 0.5, replicates = n_replicates,
                                            metrics_list = metrics_list,
                                            what = "dimensions")
    # rarefy_dim_08 <- simulation.rarefaction(remove = 0.8, replicates = n_replicates,
    #                                         metrics_list = metrics_list,
    #                                         what = "dimensions")
    # rarefy_dim_02 <- simulation.rarefaction(remove = 0.2, replicates = n_replicates,
    #                                         metrics_list = metrics_list,
    #                                         what = "dimensions")

    ## Save simulations
    save.results(rarefy_dim_05)
    # save.results(rarefy_dim_08)
    # save.results(rarefy_dim_02)
} else {
    ## Load results
    rarefy_dim_05 <- load.results("rarefy_dim_05")
    # rarefy_dim_08 <- load.results("rarefy_dim_08")
    # rarefy_dim_02 <- load.results("rarefy_dim_02")
}

## Summarise the data
rarefy_dim_05 <- lapply(rarefy_dim_05, summarise.metrics)
# rarefy_dim_08 <- lapply(rarefy_dim_08, summarise.metrics)
# rarefy_dim_02 <- lapply(rarefy_dim_02, summarise.metrics)
```

```{r run_rarefaction_elements, echo = FALSE, warning = FALSE, eval = FALSE, results = 'hide'}
#, results = 'hide'}
## Toggle data generation or not
generate_data <- FALSE

if(generate_data) {
    ## Bundle simulations
    rarefy_ele_05 <- simulation.rarefaction(remove = 0.5, replicates = n_replicates,
                                            metrics_list = metrics_list,
                                            what = "elements")
    rarefy_ele_08 <- simulation.rarefaction(remove = 0.8, replicates = n_replicates,
                                            metrics_list = metrics_list,
                                            what = "elements")
    rarefy_ele_02 <- simulation.rarefaction(remove = 0.2, replicates = n_replicates,
                                            metrics_list = metrics_list,
                                            what = "elements")

    ## Save simulations
    save.results(rarefy_ele_05)
    save.results(rarefy_ele_08)
    save.results(rarefy_ele_02)
} else {
    ## Load results
    rarefy_ele_05 <- load.results("rarefy_ele_05")
    rarefy_ele_08 <- load.results("rarefy_ele_08")
    rarefy_ele_02 <- load.results("rarefy_ele_02")
}

## Summarise the data
rarefy_ele_05 <- lapply(rarefy_ele_05, summarise.metrics)
rarefy_ele_08 <- lapply(rarefy_ele_08, summarise.metrics)
rarefy_ele_02 <- lapply(rarefy_ele_02, summarise.metrics)

```

## Plotting data data

### Results per space (remove = 50%)
```{r fig_results_05, echo = FALSE, fig.height = 8, fig.width = 10}
## Plotting the results for 0.5
plot.metrics(remove_05, col = default.palette, remove = 0.5, defaults = defaults,
    text.in.cell = TRUE)
```

### Results per space (remove = 20%)
```{r fig_results_08, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.8
plot.metrics(remove_08, col = default.palette, remove = 0.8, defaults = defaults,
    text.in.cell = TRUE)
```

### Results per space (remove = 80%)
```{r fig_results_02, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.2
plot.metrics(remove_02, col = default.palette, remove = 0.2, defaults = defaults,
    text.in.cell = TRUE)
```

### Dimensions rarefied results for PCO space (remove = 50%)
```{r fig_results_05_rare, echo = FALSE, fig.height = 8, fig.width = 10}
## Plotting the results for 0.5
plot.metrics(rarefy_dim_05, col = heat.colors, remove = 0.5, defaults = defaults,
    text.in.cell = TRUE)
```

### Dimensions rarefied results for PCO space (remove = 20%)
```{r fig_results_08_rare, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.8
plot.metrics(rarefy_dim_08, col = heat.colors, remove = 0.8, defaults = defaults,
    text.in.cell = TRUE)
```

### Dimensions rarefied results for PCO space (remove = 80%)
```{r fig_results_02_rare, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.2
plot.metrics(rarefy_dim_02, col = heat.colors, remove = 0.2, defaults = defaults,
    text.in.cell = TRUE)
```

<!-- ### Elements rarefied results for PCO space (remove = 50%) -->

```{r fig_results_05_dim, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.5
plot.metrics(rarefy_ele_05, col = heat.colors, remove = 0.5, defaults = defaults,
    text.in.cell = TRUE)
```

<!-- ### Elements rarefied results for PCO space (remove = 20%) -->

```{r fig_results_08_dim, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.8
plot.metrics(rarefy_ele_08, col = heat.colors, remove = 0.8, defaults = defaults,
    text.in.cell = TRUE)
```

<!-- ### Elements rarefied results for PCO space (remove = 80%) -->

```{r fig_results_02_dim, echo = FALSE, fig.height = 8, fig.width = 10, eval = FALSE}
## Plotting the results for 0.2
plot.metrics(rarefy_ele_02, col = heat.colors, remove = 0.2, defaults = defaults,
    text.in.cell = TRUE)
```
