---
  title: "Shifting spaces: how to summarise multidimensional spaces occupancy?"
author: "Thomas Guillerme, Mark Puttick, Vera Weisbecker"
bibliography: references.bib
date: "`r Sys.Date()`"
output:
  html_document:
  fig_width: 6
fig_height: 6
keep_tex: true
self_contained: false
line_numbers: true
fig_caption: true
---
  
  
```{r header, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
## Repeatability note:
## This whole paper is entirely reproducible and compiles as a single document. The data, figures
## and tables are all generated through the code snippets in this file. Note that the first time
## that this paper is compiled, it will generate the data which will take substantial time
## as the shiftingspace_supplementary.Rmd file completion takes ~30 minutes. Subsequent compilations
## will be much faster!

## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerm/moms"); library(moms)

## Create data directory
if(!dir.exists("../data")) {
    dir.create(path="../data", showWarnings = FALSE)
}

## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues = seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[1:n]
}

## Checking whether the data exists
if(length(grep("remove_05.Rda", list.files("../data/"))) == 0 &&
    length(grep("rarefy_dim_05.Rda", list.files("../data/"))) == 0) {
    ## Run The supplementary material
    rmarkdown::render("shiftingspace_supplementary.Rmd", "html_document") #html_document()
}
```

# Abstract

<!-- Eco/Evo option: -->
<!-- Trait multidimensional analysis are now a common tool-kit in ecology and evolution.
Such analysis are based on a multidimensional where each dimension represent a trait (or a transformation thereof) thus summarising all observed and potential trait combinations in a multidimensional trait-space (e.g. a morphospace or an ecospace).
To quantify the overall occupation of such trait-spaces, researchers apply one or more metrics that summarise the way in which organisms inhabit that 'trait-space'.
Using these metrics, researchers have investigated at how trait space occupancy changes through time, in relation to other groups of organisms or in response to global environmental changes (e.g. global warming events or mass extinctions).

Although few well defined space occupancy metrics are used in ecology, we lack a general description of the majority of those metrics, especially in macroevolution (often called disparity metrics).
Therefore the mathematical and biological meaning of such trait-space occupancy metrics can be bague.
This as consequences when interpreting changes in trait-space occupancy and linking them to biological phenomenon.
For example, does a reduction in volume in trait-space corresponds to episodes of extinction or changes in traits distribution? -->

<!-- Evo only option: -->
Analysis of the diversity of morphological shapes (aka disparity analysis) are now a common toolkit in macroevolution.
Disparity analysis are based on the analysis of a morphospace which is a multidimensional space summarising all real or potential shapes.
To quantify the overall occupation of morphospace, researchers apply one or more disparity metrics that summarise the way in which taxa inhabit 'trait space'.
Using these metrics of disparity, workers have investigated how taxa morphospace occupancy changes through time, in relation to other clades, or in response to environmental changes, such as a mass extinction.

However, we lack a general description of the majority of disparity metrics, and so the mathematical and biological meaning of disparity metrics is vague.
Thus when we interpret changes in disparity metric values we can only interpret a difference as "a change in multidimensional space occupancy". 

Seperate measures of spatial occupancy can be of particular interest in different studies.
For example, some questions might focus in changes in the volume of the morphospace whilst some other might focus on the position of a group in the morphospace or its density (or all three together).

<!-- > VW: It seems to me that the review is more important than the tool. 
> VW: So I'd be tempted to go " Here we use X method to review the properties of several commonly used disparity metrics, and introduce two new ones. 
> VW: We do this in the context of a new tool which allows easy replication of the behaviours, and also allows a fully user-controlled interface for allowing the development and testing of metrics on a diversity of datasets"
> VW: Just delete what's hyperbole :-)
> MNP - I agree with these proposed changes.
 -->
Here we propose a user friendly tool to Measure Occupancy in Multidimensional Spaces (moms) that allows to visualise numerical changes of many disparity metrics following space transformations (e.g. contraction, displacement, etc.) as well as visualising a 2D projection of the associated space changes.
This tool thus allows researchers to explore and visualise the properties of disparity metrics under various parameters.
Furthermore, we propose a thorough review of the behaviour of seven common and two new disparity metrics regarding various spacial properties (i.e. distribution, transformation, dimensionality).
As expected, we found that different metrics are more appropriated to be used in different disparity analysis.
> MNP. Rather than just observing differences, may it be more important to (i) show what different metrics measure and can we actually relate this to underlying biology, and (ii) generate null changes with parameters of interest for a researcher's clade.

# Introduction

* Disparity has been a key area of research in palaeontology, biology (Runnegar 1987; Foote 1991;...etc), and evology for the last 30 years.
* Disparity is a measure of trait variance. However, many definitions have been used in ecology (e.g. diversity of functional niches) or in evolution (e.g. diversity of morphological shapes).
* In ecology, the concept is linked to functional spaces
* Here we approach both interpretations of disparity, but with more focus on macroevolution, by defining disparity as: "multidimensional trait-space occupancy".
* In macroevolution, disparity is used to test hypothesis on how the occupancy of morphospace for one or more groups changes through time and/or space.
* In macroevolution, researchers typically form hypotheses to ask questions such as 'does disparity of one group change through time or in response to extinction?' or 'do groups A and B differ significantly in disparity?'.
* The most widely-used metrics are the sum or product of ranges and variance that measures the "spread" of a group in a morphospace (the product of ranges is the hyper-rectangular volume). However, different metrics of disparity are likely to provide more biologically-meaningful measures of change.
* For example, a group's disparity measured using the sum of variances might not change through time, but a seperate meaure can indicate a change in disparity density suggesting that the morphospace is more/less tightly packed. Additionally, the disparity of two groups measured in the Sum Of Ranges may differ even though both clades occupy the same region of the morphospace.
* Here we are going to classify disparity metrics into two broad spatial occupancy categories that can change independently: the volume, the density, and the position. All of these metrics measure the distribution of distances in trait space (e.g., Euclidean, Manhattan, or others) taht are traditionally summarised into a central tendency (e.g. the **sum** of variances, the **mean** pairwise distance), etc...
* Here we propose to test the effect of We will cite papers like this one [@disprity].



## Disparity metrics

```{r fig_metrics_types, echo = FALSE, fig.height = 3, fig.width = 9, results = 'hide', fig.cap = paste("Figure 1: different type of information captured by disparity metrics. A - Volume (e.g. sum of ranges); B - Density (e.g. average squared pairwise distances); C - Position (e.g. median distance from centroid).")}
set.seed(11)
## Plot space function (utility shortcut)
## The elements
elements <- 10

## Trait space
trait_space <- space.maker(elements, 2, distribution = rnorm)

## Graphical parameters
op <- par(mfrow = c(1,3), bty = "n")

## The volume
plot(trait_space, pch = defaults$pch, main = "A - Volume", xlab = "Trait 1", ylab = "Trait 2")
## The range lines
lines(x = range(trait_space[,1]), y = rep(mean(trait_space[,2]), 2), col = defaults$col1)
lines(x = rep(mean(trait_space[,1]), 2), y = range(trait_space[,2]), col = defaults$col1)

## The density
plot(trait_space, pch = defaults$pch, main = "B - Density", xlab = "Trait 1", ylab = "")
## Plotting the pairwise lines
pair.line <- function(line, trait_space, defaults) {
lines(x = trait_space[line,1], y = trait_space[line,2], col = defaults$col1)
}
apply(combn(1:elements, 2), 2, pair.line, trait_space, defaults)
points(trait_space, pch = defaults$pch)

## The position
plot(trait_space, pch = defaults$pch, main = "C - Position", xlab = "Trait 1", ylab = "")
## Plotting the center of the space
points(0,0, pch = 13, cex = 2)
arrows(x0 = 0, y0 = 0, x1 = trait_space[,1], y1 = trait_space[,2], code = 0, col = defaults$col1)
points(trait_space, pch = defaults$pch)

par(op)
```

Disparity metrics (or indices) can capture different aspects of the multidimensional space occupancy.
The three fundamental informations that can be captured are:

> VW: In the below can we add stuff about where metrics have resulte in debates about a group's disparity; and/or specific informations on the types of data used ; or where people are using a bunch of metrics that are already criticized (like the ones with products in them that crash to 0).

#### 1. The volume

This type of metric measures the spread of a group in the trait-space. The common metrics quantify the hyper-volume, ellipsoid volume, and product of variances or ranges. In biological terms, larger values for these metrics indicate a clade exhibits more extreme morphologies. This metric has widely been used since the advent of disparity analysis due to its implementationinf @Wills2001 in PAST or the Procrustes variance implemented in @adams2013geomorph (`geomorph::morphol.disparity`).

#### 2. The density

This type of metrics measures the "packedness" of a group in the trait-space.
In biological terms, the bigger this metric, the more different elements are from each other.
This is typically measured by pairwise distances or nearest neigbhors distances.
In macroevolution, this metric is generally used in disparity-through-time analysis based on extant taxa as implemented in @geiger2008 (`geiger::dtt`).

#### 3. The position

More rarely used, researchers summarise disparity as a measure the particular position of the group in the morphospace, either relative to another group or a fixed point in the morphospace.
The biological interpretation will thus depend on what the metrics measures the distance from.
For example if it's distance from the centre of the morphospace, the bigger the metric, the more "weird" (i.e. non-average) specimens in a group are (i.e. the centre being the average morphology in an ordination).

> VW: At the risk of sounding belligerent, the below 2 paragraphs could be more aggressive. Something like "Disparity metrics are often not designed for the question asked, but this can have serious consequences for the result. For example...[Scenario]

Of course, some metrics can be a combination of several of these three different and more.
In an ideal word, the choice of the metric should be based on the properties of changes in the morphospace of interest while taking into account the properties of the morphospace (number of elements or/and dimensions and distribution of the data on each axis).
However, it is highly unlikely that a such one-size-fits-all metric exist; rather, it is important to assess the properties of a metric relative to the specific data used.

To alleviate this problem we suggest using multiple metrics.
For example, if the macroevolutionary question of interest is to look at how the morphospace changes in response to mass extinction, it is important to capture several expected effects through different metrics (e.g. a mass extinction making the specimens with extreme characteristics go extinct - e.g. big mammals in the megafauna extinction).

> VW: but also, using your approach of multiple metrics provides opportunities for much finer-scaled hypothesis testing. So using different metrics allows the provision of space occupation information not just in one case, but several. As far as I am aware, this is new ground - e.g. you can use the kind of data you have to simulate different scenarios (e.g. to assess if an extinction was random) and then check back with your data. But it is late and have maybe just rubbished 20 years of existing disparity literature?

Finally, as different metrics will capture different aspects of the morphospace, it is important to know how such metrics will change in response to predicted or observed changes in the morphospace.
For example, when measuring the volume of a morphospace, if the morphospace space decreases randomly by 50%, do we expect a reduction of the volume of 50% as well? Or merely a reduction in density?
In this paper, we analyse the effect of such changes of the morphospace on five commonly used disparity metrics and propose two novel metrics.
We test both how these seven metrics change in response of different morphopace reduction scenarios but also how they respond to these changes depending on the space distributions and the number of dimensions.

We found no one-size-fits-all metric but provide guidance on which metric could be used in which specific scenario along with a new user friendly software with an intuitive graphical user interface (GUI) to look at the effect of changes in morphospace occupancy on user defined disparity metrics.

# Methods

To test the effect of spatial change (see Shifting space below), dimensions and dimensions distributions on the seven disparity metrics we:

1. generated 7 different spaces with variable parameters
2. transformed these spaces by removing 50% of the elements following four different scenarios
3. measured disparity on the resulted transformed spaces

Each points are described in details below.

## Generating spaces

We generated 7 spaces using the following parameters:
> VW: This needs justification or possibly a table column that cites a paper where such a scenario has been suggested/found/debated. Ideally from all the different areas of biology we can possibly reach with the paper.

| space name | size | distribution(s)                | dimensions variance | correlation |
|------------|------|--------------------------------|---------------------|-------------|
|   Uniform  |300*2 | Uniform (min = -0.5, max = 0.5)| Constant            | none        |
|   Uniform  |300*50| Uniform                        | Constant            | none        |
|   Uniform  |300*50| Uniform                        | Constant            | Random (between 0.1 and 0.9) |
|    Normal  |300*2 | Normal  (mean = 0, sd = 1)     | Constant            | none        |
|    Normal  |300*50| Normal                         | Constant            | none        |
|    Normal  |300*50| Normal                         | Constant            | Random (between 0.1 and 0.9) |
|   Random   |300*50| Normal, Uniform, Lognormal (meanlog = 0, sdlog = 1)| Constant            | none        |
| PCA-like   |300*50| Normal                         | Lognormal           | none        |
| PCO-like   |300*50| Normal                         | Normal              | none        |

Table 1: different simulated space distributions.

The five first spaces reflect specific distributions to understand the baseline effect of the metric on controlled variables.
The two last spaces (PCA and PCO-like) reflect spaces commonly used to measure changes in disparity in the macroevolutionary literature.

## Shifting spaces

We reduced the spaces by removing 50% of elements each time using the following algorithms

* **Randomly:** by randomly removing 50% of elements.  Figure 2A.

* **Limit:** by removing all elements with a distance from the centre (mean point) of the space lower or greater than a radius $\rho$ (where \$rho$ is selected such that 50% elements are selected).  Figure 2B.

* **Displacement:** by removing points similarly as above but using the distance from the furthest point from the centre (max point). Figure 2C.

* **Density:** by removing any pairs of point with a distance $D$ from each other where (where $D$ is selected such that 50% elements are selected). Figure 2D.

The algorithm to select $\rho$ or $D$ is described in more detail in in the [Appendix](#Appendix_algorithm_reduce).
We used mirror results for the limit, displacement and density reduction (i.e. the opposite data was also used).
For example, when increasing the **density** by removing the 50% most distant elements, we also run a transformation by removing 50% of the *closests* elements.
This resulted in seven different transformation by reducing the space: randomly, by limit (positive or negative), by displacement (positive or negative) or by density (increased or decreased).


```{r fig_reduce_space, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 2: different type of space reduction. Each panel displays 50% of the data points removed (in grey) using the following algorithm: A - randomly; B - by limit (removing data from the centre); C - by displacement (removing data from one side of the plot); and D - by density (removing data to increase or decrease the density).")}
set.seed(42)
op <- par(mfrow = c(2,2), bty = "n")
## The elements
elements <- 300
## The amount to remove
remove <- 0.5

## Trait space
trait_space <- space.maker(elements, 2, distribution = rnorm)
# ## The original space
# plot(trait_space, pch = 20, xlim = c(-3, 3), ylim = c(-3, 3),
#      main = "A - Original space", xlab = "Trait 1", ylab = "Trait 2")

## The randomly removed space
random_rm <- reduce.space(trait_space, type = "random", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, random_rm, main = "A - Random removal", defaults)

## The limit removal
limit_rm <- reduce.space(trait_space, type = "limit", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, limit_rm, main = "B - Limit", defaults)

## The displacement removal
displace_rm <- reduce.space(trait_space, type = "displacement", remove = 0.5)
## Moving the space to the upper right corner (positive)
plot.space(trait_space, displace_rm, main = "C - Displacement", defaults)

## The density removal
density_rm <- reduce.space(trait_space, type = "density", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, density_rm, main = "D - Density", defaults)

## List of removals (for plotting the tables later)
typical_removal <- list(space = trait_space,
remove = list("random" = random_rm,
              "limit(min)" = limit_rm,
              "limit(max)" = !limit_rm,
              "displace(positive)" = displace_rm,
              "displace(negative)" = !displace_rm,
              "density(high)" = density_rm,
              "density(low)" = !density_rm))
```

We repeated each of the data generation process (generating the space and reducing it) 20 times to obtain variance on the simulated parameters.

## Disparity metrics

We then measured seven different metrics on the resulting transformed spaces:

Name          | Definition                      | Captures | Origin           | `dispRity` function
:-------------|:-------------------------------:|----------|------------------|------|
Average squared distance | $\left(\sqrt{\sum_{i}^{n}{({q}_{i}-p_{i})^2}}\right)^{2}\times \frac{1}{n}$ | Density and Volume | @geiger2008 | `mean(dispRity::pairwise.dist(matrix)^2)`
Procrustes variance | $\sum_{i}^{m=dn}{M_{m}^2} \times \frac{1}{d}$ | Density and Volume | @adams2013geomorph | `sum(matrix^2)/nrow(matrix)`
Ellipsoid volume | $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ | Volume | @DonohueDim | `dispRity::ellipse.volume(matrix)`
Product of ranges | $\prod_{i}^{d}{\|max(k_i) - min(k_i)\|}$ | Volume | @Wills2001 | `prod(dispRity::ranges(matrix))`
Sum of ranges | $\sum_{i}^{d}{\|max(k_i) - min(k_i)\|}$ | Volume | @Wills2001 | `sum(dispRity::ranges(matrix))`
Product of variances | $\prod_{i}^{d}{\sigma^{2}{k_i}}$ | Volume | @Wills2001 | `prod(dispRity::variances(matrix))`
Sum of variances | $\sum_{i}^{d}{\sigma^{2}{k_i}}$ | Volume | @Wills2001 | `sum(dispRity::variances(matrix))`
Average displacement | $\frac{\sqrt{\sum_{i}^{n}{({k}_{n})^2}}}{\sqrt{\sum_{i}^{n}{({k}_{n}-Centroid_{k})^2}}}$ | Position | This paper | `mean(dispRity::displacements(matrix))`
Average nearest neighbour distance |$min\left(\sqrt{\sum_{i}^{n}{({q}_{i}-p_{i})^2}}\right)\times \frac{1}{n}$ | Density | This paper | `mean(dispRity::neighbours(matrix))`

```{r, all_metric, echo = FALSE}
## All metrics (exploration)
metrics_list <- list("av.pairwise" = function(matrix) return(mean(pairwise.dist(matrix)^2)),
                    "Procrustes" = function(matrix) return(sum(matrix^2)/nrow(matrix)),
                    "Volume" = ellipse.volume,
                    "PoR" = c(prod, ranges),
                    "SoR" = c(sum, ranges),
                    "PoV" = c(prod, variances),
                    "SoV" = c(sum, variances),
                    # "PoQ" = c(prod, quantiles),
                    # "SoQ" = c(sum, quantiles),
                    "av.displac" = c(mean, displacements),
                    "av.neighbo" = c(mean, neighbours))

metric_names <- c("Average squared distance", "Procrustes variance", "Ellipsoid volume", 
                  "Product of Range", "Sum of Ranges", "Product of Variances", "Sum of Variances",
                  "Mean displacement", "Mean nearest neighbours distance")
## Shorter vector name version (for clarity in the fable)
name <- metric_names
#"Product of Quantiles", "Sum of Quantiles", 
```

```{r, complex_metrics_example, echo = FALSE, eval = FALSE}
complex_metric <- list("med.mana.dist.centre" = list(c(median, centroids),
                       centroid = 0, method = "manhattan"))
```

## Effect of dimensionality

> VW: I understand teh below but only just and it is hard to read. I can help simplify.

Furthermore, high dimensional spaces ($>> 3$) are subject to the "curse of multidimensionality" [@cursedimensionality].
This is link to the fact that the more dimensions, the more sparse the data becomes.
In other words the probability of two points A and B overlapping in *n* dimensions is the probability of point A having the same value as point point B on dimension 1, 2, 3, etc...
Thus this probability decreases as a product of the number of dimensions.
This "curse" can make the interpretation of high dimensional data counter-intuitive.

For example, the product of range is expected to decrease with the number of dimensions, regardless of the number of elements on each dimensions.
This is counter-intuitive because:
1) first, the range of values on a dimension is bounded to the number of elements on this dimensions (i.e. increasing the number of observations on one dimension can only increase the range, never decrease it);
2) second, products are a classical example of the "curse" where multiplying a higher number of dimensions with a non-null probability of having a range of zero will irremediably result in 0 (i.e. an infinite number of positive values by 0 is still 0).


```{r fig_curse, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 3: Effect of the curse of dimensionality on different scaled disparity metrics. Metrics with containing products (ellipsoid volume, product of ranges or variances) are classically affected by the \"curse\" as they decrease more or less rapidly with an increase in dimensions. Note especially how the Product of Range actually increases and then decreases with the number of dimensions.")}
set.seed(42)
## Make a 50 dimensions space
space <- space.maker(300, 50, rnorm, scree = rev(cumsum(rep(1/50, 50))))

## Function wrappers
disparity <- function(metric, dimensions, space) {
    fast.disparity(group = 1:nrow(space), space = space, metric = metric, rare.dim = dimensions)
}
lapply.fun <- function(dimensions, metrics_list, space) {
    unlist(lapply(metrics_list, disparity, dimensions = dimensions, space = space))
}

## Curse results
curse_results <- sapply(2:50, lapply.fun, metrics_list = metrics_list, space = space)
curse_results <- apply(curse_results, 1, function(x) return(x/max(x)))

## Empty plot
hue_colors <- default.palette(ncol(curse_results))
par(bty = "n")
plot(NULL, xlim = c(2,50), ylim = c(0,1),
xlab = "Dimensions (decreasing variance)",
ylab = "Scaled disparity")
for(metric in 1:ncol(curse_results)) {
    lines(curse_results[, metric], col = hue_colors[metric])
}
legend("bottomleft", metric_names, lty = 1, col = hue_colors)
```

## Measuring the effect of space reduction, distribution and dimensionality on the metric

### Scaling disparity

To measure the effect of space reduction, distribution and dimensionality on the metric, we first scaled the metric to be relative to the non reduced space for each dimension distributions or number of dimensions.
To do so, we subtracted the observed disparity with no space reduction (i.e. base disparity) to all the disparity measurements of the reduced spaces and then divided it by the resulting maximum observed disparity.
This way, our disparity metrics where scaled between -1 and 1 with a value of 0 indicating no effect of the space reduction and $>0$ and $<0$ respectively indicating an increase or decrease in the disparity metric value.

### Measuring the effect of the metric

TODO: probably simply do a Bhattacharyya coefficient to see how much the disparity from the reduced space overlaps with the randomly reduced disparity.
High values will indicate low effect and low values, high effects.
TODO: probably flip that metric over so that low values = low effects and high values = high effects

$\text{Effect of metric} = 1 - BC(metric_{focal}, metric_{random})$

### Measuring the effect of space and dimensionality

To measure the effect of the different space distributions we run two analysis of variance: each individual disparity metric function of the different seven different spaces and each individual disparity metric function of the 10 different rarefaction levels.
The results (_p_-values) are reported in Table 3.

TODO: report the full tests

> VW:Here or perhaps in the results, I think there needs to be an explicit expectation of what "good" behaviour is for each metric. Something like an expectation - for the disparity metric to work, what should it do? E.g. If you apply the "limit" scenario, the squared distance measure is expected to go down for the black dots and up for the grey dots. 


# Results

```{r loading_results, echo = FALSE}
## Loading the results
remove_05 <- load.results("remove_05")
rarefy_dim_05 <- load.results("rarefy_dim_05")
```

```{r running_tests, echo = FALSE}
## Anova function
anova.fun <- function(data) {return(aov(glm(disparity ~ factor, data = data)))}

## Running the tests
test_space05 <- test.metrics(remove_05, test = anova.fun, scale = TRUE)
test_raref05 <- test.metrics(rarefy_dim_05, test = anova.fun, scale = TRUE)

#TODO: Generate the aov tables in the supplementary

```

```{r fable_results, fig.show='hide', echo=FALSE, fig.height=1.5, fig.width=1.5}
## Making a list of parameters for each mini plot
plot.param <- list(scaler = 0.5,
bg.col = "black",
col = c("grey", "orange", "blue"),
quantiles = c(95, 50),
cent.tend = median,
pch = 19)

## Looping through each mini plot for every metric
for(metric in 1:length(metrics_list)) {
    generate.fable.plot(data = remove_05, metric = metric, what = "limits", plot.param = plot.param)
    generate.fable.plot(data = remove_05, metric = metric, what = "displa", plot.param = plot.param)
    generate.fable.plot(data = remove_05, metric = metric, what = "densit", plot.param = plot.param)
}

```

> VW:I am stuck here. YOu show 4 scenarios but only 3 plots per metric and the names of the scenarios does not match. I want to make sure I understand this right.

Table 3: Results of the effect of space reduction, space dimensions distributions and dimensions number of the different disparity metrics.
Figures: the dots represent the median disparity values, the solid and dashed line respectively the 50% and 95% confidence intervals. Results in grey are the random 50% reduction, in orange the negative reductions (i.e. discarding the black dots in Figure 2) and in blue the positive reductions (i.e. discarding the grey dots in Figure 2). _p_-values for distribution effect and dimensions effect represents respectively the effect of the ANOVAs disparity ~ distributions and disparity ~ dimensions (Signif. codes:  0 '\*\*\*' 0.001 '\*\*' 0.01 '\*' 0.05 '.' 0.1 ' ' 1).

Metric      | Volume change  | Position change | Density change | Distribution effect | Dimensions effect
:-----------|----------------|-----------------|----------------|---------------------|-------------------
`r name[1]` |`r plot.id(1)`  |`r plot.id(2)`   |`r plot.id(3)`  | `r s.test(1, "s")`  | `r s.test(1, "r")`
`r name[2]` |`r plot.id(4)`  |`r plot.id(5)`   |`r plot.id(6)`  | `r s.test(2, "s")`  | `r s.test(2, "r")`
`r name[3]` |`r plot.id(7)`  |`r plot.id(8)`   |`r plot.id(9)`  | `r s.test(3, "s")`  | `r s.test(3, "r")`
`r name[4]` |`r plot.id(10)` |`r plot.id(11)`  |`r plot.id(12)` | `r s.test(4, "s")`  | `r s.test(4, "r")`
`r name[5]` |`r plot.id(13)` |`r plot.id(14)`  |`r plot.id(15)` | `r s.test(5, "s")`  | `r s.test(5, "r")`
`r name[6]` |`r plot.id(16)` |`r plot.id(17)`  |`r plot.id(18)` | `r s.test(6, "s")`  | `r s.test(6, "r")`
`r name[7]` |`r plot.id(19)` |`r plot.id(20)`  |`r plot.id(21)` | `r s.test(7, "s")`  | `r s.test(7, "r")`
`r name[8]` |`r plot.id(22)` |`r plot.id(23)`  |`r plot.id(24)` | `r s.test(8, "s")`  | `r s.test(8, "r")`
`r name[9]` |`r plot.id(25)` |`r plot.id(26)`  |`r plot.id(27)` | `r s.test(9, "s")`  | `r s.test(9, "r")`


# Discussion


* Average squared distances: all right metric for measuring density (increase/decrease picked up). Not great for volume (i.e. volume decreases when the morphospace becomes hollow). Not at all good for position (both negative/positive are picked up the same way). Affected by the distributions but not the dimensions.

* Procrustes variances: similar to the average squared distances although even worth for position (often close to random changes). Clearly not affect by the distributions and the dimensions.

* Ellispsoid volume: Not great for picking up anything, probably because of the curse of multidimensionality. This metric is clearly affected both by the distributions and the number of dimensions. This metric is probably best used with a low number of dimensions ($<10$).

* Product of ranges: Same as the ellipsoid volume.

* Sum of ranges: Picks a only negative change in volume (but not hollowness) and bat for measuring changes in position and density. This metric is clearly affected by the distributions and also by the number of dimensions (albeit less clearly).

* Product of variances: Positive decrease in volume (hollowness) is picked up. But not much signal from change in position or density. This metric is clearly affected by the distributions and the dimensions.

* Sum of variances: Similar behaviour than the Procrustes variance. Picks up great differences in volume. Bad for position on average, not bad for density. 

* Mean displacement: Picks up reduction in volume (increased metric), not great for change in position and density. This metric is clearly affected by both the distributions and the number of dimensions.

* Mean nearest neighbour distances: Picks up reduction in volume. Not great for change in position. OK for change in density. This metric is affected by the distributions but not the number of dimensions.

## Caveat of the analysis

* Data not simulated using evolutionary models (TODO: maybe TODO?)
* Response to reduction is often more specific to distribution/dimensions/reduction for each metric and these fine differences are not displayed here.
*

# Conclusion

There are non one-size-fits-all disparity metric.
It is important to use the disparity metric that will best capture the problem at hand (e.g. if interested in the position of the morphospace, don't use the product of variances).

Furthermore, all these different metrics are "disparity metrics" even though they represent widely different aspects.
We thus suggest that future disparity analysis authors abstain from using the term "disparity" to designate the "disparity metric" but rather use "position in the morphospace", or, directly the name of the used metric to avoid confusion.


# Acknowledgements

Money from the ARC.
ARC DP170103227, FT180100634



# References





# Supplementary material

## Algorithm for selecting the parameters to reduce the space ($radius$, $displacement$, $density$) {#Appendix_algorithm_reduce}

We used a recursive algorithm for selecting the parameter that removes $P$ elements: 80%, 50% or 20%.

1. Select a random reduction parameters $R$.
2. Remove elements from the space using $R$ resulting in $P'$ removed elements.
If the remaining number of elements is to the required proportion $P$ ; exit the algorithm;
Else go to 3.
3. Get the different between the proportion of removed elements $P'$ and $P$.
If the difference is positive set the increment parameter $R$ to $R = 1.1 \times R$, then go to 2.
Else set $R = 0.9 \times R$, then go to 2.

The algorithm is implemented in the `optimise.parameter` function in [reduce.space_fun.R](@@@).
