---
title: "Shifting spaces: which disparity or dissimilarity measurement best summarise occupancy in multidimensional spaces?"
author: "Thomas Guillerme, Mark N. Puttick, Ariel E. Marcy, Vera Weisbecker"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  pdf_document:
    fig_width: 12
    fig_height: 6
---

# Supplementary material 4: full results

```{r header_simulation, echo = FALSE}
## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerm/moms"); library(moms)

## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues = seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[1:n]
}
```

```{r compilation_html, echo = FALSE, eval = FALSE}
## Changing defaults
body(plot.id)[[2]] <- substitute(type <- ".png")
```

# Simulation data

This section contains all the results and calculations (and plots them).
The whole analysis takes approximatively 30 minutes to run (single core 2.2 GHz).

## Measurements analysed


```{r all_metrics_list, echo = FALSE}
## Loading the list of metrics
source("list.of.metrics.R")
## General size measures
all_metrics <- list("av.pairwise" = function(matrix) return(mean(pairwise.dist(matrix)^2, na.rm = TRUE)),
                    "av.pair.ma"  = function(matrix) return(mean(pairwise.dist(matrix, method = "manhattan"), na.rm = TRUE)),
                    "med.pairwise"= function(matrix) return(median(pairwise.dist(matrix), na.rm = TRUE)),
                    "med.pair.ma" = function(matrix) return(median(pairwise.dist(matrix, method = "manhattan"), na.rm = TRUE)),
                    "Procrustes"  = function(matrix) return(sum(matrix^2)/nrow(matrix)),
                    "med.cent"    = c(median, centroids),
                    "med.cent.man"= function(matrix) return(median(centroids(matrix, method = "manhattan"))),
                    "sum.var"     = c(sum, variances),
                    "sum.ranges"  = c(sum, ranges),
                    "sum.quantil" = c(sum, quantiles),
                    "prod.var"    = c(prod, variances),
                    "prod.ranges" = c(prod, ranges),
                    "prod.quantil"= c(prod, quantiles),
                    "ellips.vol"  = ellipse.volume,
                    "n.ball.vol"  = n.ball.volume,
                    "span.tree"   = c(mean, span.tree.length),
                    "ave.neigh"   = c(mean, neighbours),
                    "ave.neig.man"= function(matrix) return(mean(neighbours(matrix, method = "manhattan"))),
                    "func.div"    = func.div,
                    "func.eve"    = func.eve,
                    "func.disp"   = func.disp,
                    "av.displa"   = c(mean, displacements),
                    "av.displ.man"= function(matrix) return(mean(displacements(matrix, method = "manhattan"))),
                    "med.center"  = function(matrix) return(median(centroids(matrix, centroid = 0))),
                    "med.cent.ma" = function(matrix) return(median(centroids(matrix, centroid = 0, method = "manhattan")))
                    )

all_metrics_names <- c("Average squared pairwise distance (dtt::dtt)",
                       "Average pairwise manhattan distance (dtt::dtt)",
                       "Median pairwise distance",
                       "Median pairwise manhattan distance",
                       "Procrustes variances (geomorph::morpho.disparity)",
                       "Median distance from centroid",
                       "Median manhattan distance from centroid",
                       "Sum of variances",
                       "Sum of ranges",
                       "Sum of quantiles",
                       "Product of variances",
                       "Product of ranges",
                       "Product of quantiles",
                       "Ellipsoid volume",
                       "nBall volume",
                       "Minimum spanning tree average length",
                       "Average minimum neighbours distance",
                       "Average minimum neighbours manhattan distance",
                       "Function diversity",
                       "Functional evenness",
                       "Functional dispersion",
                       "Average displacement",
                       "Average manhattan displacement",
                       "Median distance from centre",
                       "Median manhattan distance from centre"
                       )


```

The measurements analysed are the `r paste(all_metrics_names, collapse = ", ")`.

## Running the simulations for the 25 metrics

```{r start_simulation, echo = FALSE}
## Set the overall number of replicates
n_replicates <- 20

## Set the overall number of elements
#elements <- function() sample(20:200) #TG: possibility to use a variable number of elements
elements <- function() 200

## Simulation seed
set.seed(42)
```

```{r generate_spaces_simulation, echo = FALSE, warning = FALSE, results = 'hide', eval = FALSE}
## List of distributions
distributions_list <- list(
    "unifor" = list(distribution = runif, arguments = list(list("min" = -0.5, "max" = 0.5))),
    "normal" = list(distribution = rnorm),
    "random" = list(distribution = "random"),
    "unicor" = list(distribution = runif, arguments = list(list("min" = -0.5, "max" = 0.5)),
                    cor.matrix = "random"),
    "norcor" = list(distribution = rnorm, cor.matrix = "random"),
    "pcalik" = list(distribution = rnorm, scree = "lognormal"),
    "pcolik" = list(distribution = rnorm, scree = "normal")
    )

## Generate all 13 spaces
all_spaces <- list(uniform3 = space.simulation(distributions_list$unifor, dimensions = 3,
                                               elements = elements(), replicates = n_replicates),
                   uniform15 = space.simulation(distributions_list$unifor, dimensions = 15,
                                               elements = elements(), replicates = n_replicates),
                   uniform50 = space.simulation(distributions_list$unifor, dimensions = 50,
                                               elements = elements(), replicates = n_replicates),
                   uniform100 = space.simulation(distributions_list$unifor, dimensions = 100,
                                               elements = elements(), replicates = n_replicates),
                   uniform150 = space.simulation(distributions_list$unifor, dimensions = 150,
                                               elements = elements(), replicates = n_replicates),
                   uniformc50 = space.simulation(distributions_list$unicor, dimensions = 50,
                                               elements = elements(), replicates = n_replicates),
                   normal3 = space.simulation(distributions_list$normal, dimensions = 3,
                                               elements = elements(), replicates = n_replicates),
                   normal15 = space.simulation(distributions_list$normal, dimensions = 15,
                                               elements = elements(), replicates = n_replicates),
                   normal50 = space.simulation(distributions_list$normal, dimensions = 50,
                                               elements = elements(), replicates = n_replicates),
                   uniform100 = space.simulation(distributions_list$unifor, dimensions = 100,
                                               elements = elements(), replicates = n_replicates),
                   normal150 = space.simulation(distributions_list$normal, dimensions = 150,
                                               elements = elements(), replicates = n_replicates),
                   normalc50 = space.simulation(distributions_list$norcor, dimensions = 50,
                                               elements = elements(), replicates = n_replicates),
                   random50 = space.simulation(distributions_list$random, dimensions = 50,
                                               elements = elements(), replicates = n_replicates),
                   pca_like = space.simulation(distributions_list$pcalik, dimensions = 50,
                                               elements = elements(), replicates = n_replicates),
                   pco_like = space.simulation(distributions_list$pcolik, dimensions = 50,
                                               elements = elements(), replicates = n_replicates))

#TG: Warnings are due to inexact correlations
```

```{r shift_spaces_simulation, echo = FALSE, warning = FALSE, results = 'hide', eval = FALSE}
## Shifting each space
shift_groups_02 <- lapply(all_spaces, lapply, shift.group.simulation, remove = 0.2)
shift_groups_05 <- lapply(all_spaces, lapply, shift.group.simulation, remove = 0.5)
shift_groups_08 <- lapply(all_spaces, lapply, shift.group.simulation, remove = 0.8)
```

```{r get_disparity_simulation, echo = FALSE, warning = FALSE, results = 'hide', eval = FALSE}
## Measuring disparity
all_metrics_remove_02 <- mapply(metrics.simulation, all_spaces, shift_groups_02,
                        MoreArgs = list(metrics = all_metrics, rare.dim = NULL, verbose = TRUE),
                        SIMPLIFY = FALSE)
all_metrics_remove_05 <- mapply(metrics.simulation, all_spaces, shift_groups_05,
                        MoreArgs = list(metrics = all_metrics, rare.dim = NULL, verbose = TRUE),
                        SIMPLIFY = FALSE)
all_metrics_remove_08 <- mapply(metrics.simulation, all_spaces, shift_groups_08,
                        MoreArgs = list(metrics = all_metrics, rare.dim = NULL, verbose = TRUE),
                        SIMPLIFY = FALSE)

## Saving the results
save.results(all_metrics_remove_02)
save.results(all_metrics_remove_05)
save.results(all_metrics_remove_08)
```

## Running the empirical data for the 25 metrics

```{r running_empirical, echo = FALSE, eval = FALSE}
data(demo_data)
## Loading the data
trait_spaces <- demo_data

## Measuring space occupancy for all spaces and all metrics
apply.disparity <- function(space, metrics) {
    lapply(metrics, function(metric, space) dispRity(boot.matrix(space, bootstraps = 500), metric),
           space)
}
all_metrics_empirical_results <- lapply(trait_spaces, apply.disparity, all_metrics)

## Results list
save.results(all_metrics_empirical_results)
```


## Load all the data


```{r, loading_data}
## Load all the data
all_metrics_remove_02 <- load.results("all_metrics_remove_02")
all_metrics_remove_05 <- load.results("all_metrics_remove_05")
all_metrics_remove_08 <- load.results("all_metrics_remove_08")
all_metrics_empirical_results <- load.results("all_metrics_empirical_results")
```


## Measurements pairwise comparisons

```{r, pairwise_comparisons, fig.height = 12}
## getting the results to loop through (independent of the removal)
results_pairwise <- all_metrics_remove_02

## Plotting the pairwise results
for(metric_ID in 1:length(all_metrics_names)) {
  pairwise.plot(results_pairwise, scale = TRUE, type = "base", plot = metric_ID)
  cat(paste0(all_metrics_names[metric_ID], "\n"))
}
```

The measurements analysed are the `r paste(paste(names(all_metrics), all_metrics_names, sep = ": "), collapse = "; ")`.


## Space shifting all measures


```{r running_anovas, echo = FALSE}
## Anova function
anova.fun <- function(data) {return(aov(glm(disparity ~ factor, data = data)))}

## Running the tests
all_test_02  <- test.simulation(all_metrics_remove_02, test = anova.fun, scale = TRUE)
all_test_05  <- test.simulation(all_metrics_remove_05, test = anova.fun, scale = TRUE)
all_test_08  <- test.simulation(all_metrics_remove_08, test = anova.fun, scale = TRUE)

all_dim_test_02 <- test.simulation(all_metrics_remove_02, test = anova.fun, scale = TRUE,
                      factors = c("uniform3", "uniform15", "uniform50", "uniform100", "uniform150",
                                  "normal3", "normal15", "normal50", "normal100", "normal150"))
all_dim_test_05 <- test.simulation(all_metrics_remove_05, test = anova.fun, scale = TRUE,
                      factors = c("uniform3", "uniform15", "uniform50", "uniform100", "uniform150",
                                  "normal3", "normal15", "normal50", "normal100", "normal150"))
all_dim_test_08 <- test.simulation(all_metrics_remove_08, test = anova.fun, scale = TRUE,
                      factors = c("uniform3", "uniform15", "uniform50", "uniform100", "uniform150",
                                  "normal3", "normal15", "normal50", "normal100", "normal150"))

space_test_02 <- test.simulation(all_metrics_remove_02, test = anova.fun, scale = TRUE,
                              factors = c("uniform50", "uniform50c", "normal50", "normal50c",
                                          "random50", "pca_like", "pco_like"))
space_test_05 <- test.simulation(all_metrics_remove_05, test = anova.fun, scale = TRUE,
                              factors = c("uniform50", "uniform50c", "normal50", "normal50c",
                                          "random50", "pca_like", "pco_like"))
space_test_08 <- test.simulation(all_metrics_remove_08, test = anova.fun, scale = TRUE,
                              factors = c("uniform50", "uniform50c", "normal50", "normal50c",
                                          "random50", "pca_like", "pco_like"))
```

```{r fable_results options}
## The metrics names (shortened vector)
name <- all_metrics_names

## Making a list of parameters for each mini plot
plot.param <- list(scaler = 3,
                   bg.col = "black",
                   col = c("grey", "orange", "blue"),
                   quantiles = c(95, 50),
                   cent.tend = median,
                   pch = 19,
                   metric.max = length(all_metrics),
                   cex = 2)
```

```{r fable_results_all_metrics_02, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Looping through each mini plot for every metric
for(metric in 1:length(all_metrics)) {
    generate.fable.plot(data = all_metrics_remove_02, metric = metric, what = "limits", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = all_metrics_remove_02, metric = metric, what = "densit", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = all_metrics_remove_02, metric = metric, what = "displa", plot.param = plot.param, overlap = TRUE)
}
```

```{r fable_results_all_metrics_05, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Looping through each mini plot for every metric
for(metric in 1:length(all_metrics)) {
    generate.fable.plot(data = all_metrics_remove_05, metric = metric, what = "limits", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = all_metrics_remove_05, metric = metric, what = "densit", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = all_metrics_remove_05, metric = metric, what = "displa", plot.param = plot.param, overlap = TRUE)
}
```

```{r fable_results_all_metrics_08, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Looping through each mini plot for every metric
for(metric in 1:length(all_metrics)) {
    generate.fable.plot(data = all_metrics_remove_08, metric = metric, what = "limits", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = all_metrics_remove_08, metric = metric, what = "densit", plot.param = plot.param, overlap = TRUE)
    generate.fable.plot(data = all_metrics_remove_08, metric = metric, what = "displa", plot.param = plot.param, overlap = TRUE)
}
```

```{r, echo = FALSE, eval = FALSE}
## Function for printing the table in the R console
print.fable <- function(n_metrics, byrow, ncol, test) {
  ## Make the plot.id table 
  ids <- matrix(1:(n_metrics*ncol), ncol = ncol, byrow = byrow)

  for(one_metric in 1:n_metrics) {
    ## Print all the rows one by one
    text <- c(paste0("`r name[", one_metric, "]`"), paste0("`r plot.id(", ids[one_metric, ], ")`"))
    ## Add some tests?
    if(test) {
      text <- c(text, paste0("`r s.test(", one_metric, ", \"s\", res)`"), paste0("`r s.test(", one_metric, ", \"r\", res)`"))
    }
    ## Print the line
    cat(paste(text, collapse = " | "))
    cat("|\n")
  }
}
## Getting the fable to copy paste under the table header below.
print.fable(length(all_metrics), byrow = TRUE, ncol = 3, test = TRUE)
```

# Table 1: All measures results 20% removal

```{r fable_all_metrics_02, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[3]] <- substitute(chain <- "fable_results_all_metrics_02")
res <- list("s" = space_test_02, "r" = all_dim_test_02)
```

Metric      | Size change  | Density change | Position change | Distribution effect | Dimensions effect |
:-----------|----------------|-----------------|----------------|---------------------|-------------------|
`r name[1]` | `r plot.id(1)` | `r plot.id(2)` | `r plot.id(3)` | `r s.test(1, "s", res)` | `r s.test(1, "r", res)`|
`r name[2]` | `r plot.id(4)` | `r plot.id(5)` | `r plot.id(6)` | `r s.test(2, "s", res)` | `r s.test(2, "r", res)`|
`r name[3]` | `r plot.id(7)` | `r plot.id(8)` | `r plot.id(9)` | `r s.test(3, "s", res)` | `r s.test(3, "r", res)`|
`r name[4]` | `r plot.id(10)` | `r plot.id(11)` | `r plot.id(12)` | `r s.test(4, "s", res)` | `r s.test(4, "r", res)`|
`r name[5]` | `r plot.id(13)` | `r plot.id(14)` | `r plot.id(15)` | `r s.test(5, "s", res)` | `r s.test(5, "r", res)`|
`r name[6]` | `r plot.id(16)` | `r plot.id(17)` | `r plot.id(18)` | `r s.test(6, "s", res)` | `r s.test(6, "r", res)`|
`r name[7]` | `r plot.id(19)` | `r plot.id(20)` | `r plot.id(21)` | `r s.test(7, "s", res)` | `r s.test(7, "r", res)`|
`r name[8]` | `r plot.id(22)` | `r plot.id(23)` | `r plot.id(24)` | `r s.test(8, "s", res)` | `r s.test(8, "r", res)`|
`r name[9]` | `r plot.id(25)` | `r plot.id(26)` | `r plot.id(27)` | `r s.test(9, "s", res)` | `r s.test(9, "r", res)`|
`r name[10]` | `r plot.id(28)` | `r plot.id(29)` | `r plot.id(30)` | `r s.test(10, "s", res)` | `r s.test(10, "r", res)`|
`r name[11]` | `r plot.id(31)` | `r plot.id(32)` | `r plot.id(33)` | `r s.test(11, "s", res)` | `r s.test(11, "r", res)`|
`r name[12]` | `r plot.id(34)` | `r plot.id(35)` | `r plot.id(36)` | `r s.test(12, "s", res)` | `r s.test(12, "r", res)`|
`r name[13]` | `r plot.id(37)` | `r plot.id(38)` | `r plot.id(39)` | `r s.test(13, "s", res)` | `r s.test(13, "r", res)`|
`r name[14]` | `r plot.id(40)` | `r plot.id(41)` | `r plot.id(42)` | `r s.test(14, "s", res)` | `r s.test(14, "r", res)`|
`r name[15]` | `r plot.id(43)` | `r plot.id(44)` | `r plot.id(45)` | `r s.test(15, "s", res)` | `r s.test(15, "r", res)`|
`r name[16]` | `r plot.id(46)` | `r plot.id(47)` | `r plot.id(48)` | `r s.test(16, "s", res)` | `r s.test(16, "r", res)`|
`r name[17]` | `r plot.id(49)` | `r plot.id(50)` | `r plot.id(51)` | `r s.test(17, "s", res)` | `r s.test(17, "r", res)`|
`r name[18]` | `r plot.id(52)` | `r plot.id(53)` | `r plot.id(54)` | `r s.test(18, "s", res)` | `r s.test(18, "r", res)`|
`r name[19]` | `r plot.id(55)` | `r plot.id(56)` | `r plot.id(57)` | `r s.test(19, "s", res)` | `r s.test(19, "r", res)`|
`r name[20]` | `r plot.id(58)` | `r plot.id(59)` | `r plot.id(60)` | `r s.test(20, "s", res)` | `r s.test(20, "r", res)`|
`r name[21]` | `r plot.id(61)` | `r plot.id(62)` | `r plot.id(63)` | `r s.test(21, "s", res)` | `r s.test(21, "r", res)`|
`r name[22]` | `r plot.id(64)` | `r plot.id(65)` | `r plot.id(66)` | `r s.test(22, "s", res)` | `r s.test(22, "r", res)`|
`r name[23]` | `r plot.id(67)` | `r plot.id(68)` | `r plot.id(69)` | `r s.test(23, "s", res)` | `r s.test(23, "r", res)`|
`r name[24]` | `r plot.id(70)` | `r plot.id(71)` | `r plot.id(72)` | `r s.test(24, "s", res)` | `r s.test(24, "r", res)`|
`r name[25]` | `r plot.id(73)` | `r plot.id(74)` | `r plot.id(75)` | `r s.test(25, "s", res)` | `r s.test(25, "r", res)`|


# Table 2: All measures results 50% removal

```{r fable_all_metrics_05, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[3]] <- substitute(chain <- "fable_results_all_metrics_05")
res <- list("s" = space_test_05, "r" = all_dim_test_05)
```

Measurements      | Size change  | Density change | Position change | Distribution effect | Dimensions effect |
:-----------|----------------|-----------------|----------------|---------------------|-------------------|
`r name[1]` | `r plot.id(1)` | `r plot.id(2)` | `r plot.id(3)` | `r s.test(1, "s", res)` | `r s.test(1, "r", res)`|
`r name[2]` | `r plot.id(4)` | `r plot.id(5)` | `r plot.id(6)` | `r s.test(2, "s", res)` | `r s.test(2, "r", res)`|
`r name[3]` | `r plot.id(7)` | `r plot.id(8)` | `r plot.id(9)` | `r s.test(3, "s", res)` | `r s.test(3, "r", res)`|
`r name[4]` | `r plot.id(10)` | `r plot.id(11)` | `r plot.id(12)` | `r s.test(4, "s", res)` | `r s.test(4, "r", res)`|
`r name[5]` | `r plot.id(13)` | `r plot.id(14)` | `r plot.id(15)` | `r s.test(5, "s", res)` | `r s.test(5, "r", res)`|
`r name[6]` | `r plot.id(16)` | `r plot.id(17)` | `r plot.id(18)` | `r s.test(6, "s", res)` | `r s.test(6, "r", res)`|
`r name[7]` | `r plot.id(19)` | `r plot.id(20)` | `r plot.id(21)` | `r s.test(7, "s", res)` | `r s.test(7, "r", res)`|
`r name[8]` | `r plot.id(22)` | `r plot.id(23)` | `r plot.id(24)` | `r s.test(8, "s", res)` | `r s.test(8, "r", res)`|
`r name[9]` | `r plot.id(25)` | `r plot.id(26)` | `r plot.id(27)` | `r s.test(9, "s", res)` | `r s.test(9, "r", res)`|
`r name[10]` | `r plot.id(28)` | `r plot.id(29)` | `r plot.id(30)` | `r s.test(10, "s", res)` | `r s.test(10, "r", res)`|
`r name[11]` | `r plot.id(31)` | `r plot.id(32)` | `r plot.id(33)` | `r s.test(11, "s", res)` | `r s.test(11, "r", res)`|
`r name[12]` | `r plot.id(34)` | `r plot.id(35)` | `r plot.id(36)` | `r s.test(12, "s", res)` | `r s.test(12, "r", res)`|
`r name[13]` | `r plot.id(37)` | `r plot.id(38)` | `r plot.id(39)` | `r s.test(13, "s", res)` | `r s.test(13, "r", res)`|
`r name[14]` | `r plot.id(40)` | `r plot.id(41)` | `r plot.id(42)` | `r s.test(14, "s", res)` | `r s.test(14, "r", res)`|
`r name[15]` | `r plot.id(43)` | `r plot.id(44)` | `r plot.id(45)` | `r s.test(15, "s", res)` | `r s.test(15, "r", res)`|
`r name[16]` | `r plot.id(46)` | `r plot.id(47)` | `r plot.id(48)` | `r s.test(16, "s", res)` | `r s.test(16, "r", res)`|
`r name[17]` | `r plot.id(49)` | `r plot.id(50)` | `r plot.id(51)` | `r s.test(17, "s", res)` | `r s.test(17, "r", res)`|
`r name[18]` | `r plot.id(52)` | `r plot.id(53)` | `r plot.id(54)` | `r s.test(18, "s", res)` | `r s.test(18, "r", res)`|
`r name[19]` | `r plot.id(55)` | `r plot.id(56)` | `r plot.id(57)` | `r s.test(19, "s", res)` | `r s.test(19, "r", res)`|
`r name[20]` | `r plot.id(58)` | `r plot.id(59)` | `r plot.id(60)` | `r s.test(20, "s", res)` | `r s.test(20, "r", res)`|
`r name[21]` | `r plot.id(61)` | `r plot.id(62)` | `r plot.id(63)` | `r s.test(21, "s", res)` | `r s.test(21, "r", res)`|
`r name[22]` | `r plot.id(64)` | `r plot.id(65)` | `r plot.id(66)` | `r s.test(22, "s", res)` | `r s.test(22, "r", res)`|
`r name[23]` | `r plot.id(67)` | `r plot.id(68)` | `r plot.id(69)` | `r s.test(23, "s", res)` | `r s.test(23, "r", res)`|
`r name[24]` | `r plot.id(70)` | `r plot.id(71)` | `r plot.id(72)` | `r s.test(24, "s", res)` | `r s.test(24, "r", res)`|
`r name[25]` | `r plot.id(73)` | `r plot.id(74)` | `r plot.id(75)` | `r s.test(25, "s", res)` | `r s.test(25, "r", res)`|

# Table 3: All measures results 80% removal

```{r fable_all_metrics_08, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[3]] <- substitute(chain <- "fable_results_all_metrics_08")
res <- list("s" = space_test_08, "r" = all_dim_test_08)
```

Measurements      | Size change  | Density change | Position change | Distribution effect | Dimensions effect |
:-----------|----------------|-----------------|----------------|---------------------|-------------------|
`r name[1]` | `r plot.id(1)` | `r plot.id(2)` | `r plot.id(3)` | `r s.test(1, "s", res)` | `r s.test(1, "r", res)`|
`r name[2]` | `r plot.id(4)` | `r plot.id(5)` | `r plot.id(6)` | `r s.test(2, "s", res)` | `r s.test(2, "r", res)`|
`r name[3]` | `r plot.id(7)` | `r plot.id(8)` | `r plot.id(9)` | `r s.test(3, "s", res)` | `r s.test(3, "r", res)`|
`r name[4]` | `r plot.id(10)` | `r plot.id(11)` | `r plot.id(12)` | `r s.test(4, "s", res)` | `r s.test(4, "r", res)`|
`r name[5]` | `r plot.id(13)` | `r plot.id(14)` | `r plot.id(15)` | `r s.test(5, "s", res)` | `r s.test(5, "r", res)`|
`r name[6]` | `r plot.id(16)` | `r plot.id(17)` | `r plot.id(18)` | `r s.test(6, "s", res)` | `r s.test(6, "r", res)`|
`r name[7]` | `r plot.id(19)` | `r plot.id(20)` | `r plot.id(21)` | `r s.test(7, "s", res)` | `r s.test(7, "r", res)`|
`r name[8]` | `r plot.id(22)` | `r plot.id(23)` | `r plot.id(24)` | `r s.test(8, "s", res)` | `r s.test(8, "r", res)`|
`r name[9]` | `r plot.id(25)` | `r plot.id(26)` | `r plot.id(27)` | `r s.test(9, "s", res)` | `r s.test(9, "r", res)`|
`r name[10]` | `r plot.id(28)` | `r plot.id(29)` | `r plot.id(30)` | `r s.test(10, "s", res)` | `r s.test(10, "r", res)`|
`r name[11]` | `r plot.id(31)` | `r plot.id(32)` | `r plot.id(33)` | `r s.test(11, "s", res)` | `r s.test(11, "r", res)`|
`r name[12]` | `r plot.id(34)` | `r plot.id(35)` | `r plot.id(36)` | `r s.test(12, "s", res)` | `r s.test(12, "r", res)`|
`r name[13]` | `r plot.id(37)` | `r plot.id(38)` | `r plot.id(39)` | `r s.test(13, "s", res)` | `r s.test(13, "r", res)`|
`r name[14]` | `r plot.id(40)` | `r plot.id(41)` | `r plot.id(42)` | `r s.test(14, "s", res)` | `r s.test(14, "r", res)`|
`r name[15]` | `r plot.id(43)` | `r plot.id(44)` | `r plot.id(45)` | `r s.test(15, "s", res)` | `r s.test(15, "r", res)`|
`r name[16]` | `r plot.id(46)` | `r plot.id(47)` | `r plot.id(48)` | `r s.test(16, "s", res)` | `r s.test(16, "r", res)`|
`r name[17]` | `r plot.id(49)` | `r plot.id(50)` | `r plot.id(51)` | `r s.test(17, "s", res)` | `r s.test(17, "r", res)`|
`r name[18]` | `r plot.id(52)` | `r plot.id(53)` | `r plot.id(54)` | `r s.test(18, "s", res)` | `r s.test(18, "r", res)`|
`r name[19]` | `r plot.id(55)` | `r plot.id(56)` | `r plot.id(57)` | `r s.test(19, "s", res)` | `r s.test(19, "r", res)`|
`r name[20]` | `r plot.id(58)` | `r plot.id(59)` | `r plot.id(60)` | `r s.test(20, "s", res)` | `r s.test(20, "r", res)`|
`r name[21]` | `r plot.id(61)` | `r plot.id(62)` | `r plot.id(63)` | `r s.test(21, "s", res)` | `r s.test(21, "r", res)`|
`r name[22]` | `r plot.id(64)` | `r plot.id(65)` | `r plot.id(66)` | `r s.test(22, "s", res)` | `r s.test(22, "r", res)`|
`r name[23]` | `r plot.id(67)` | `r plot.id(68)` | `r plot.id(69)` | `r s.test(23, "s", res)` | `r s.test(23, "r", res)`|
`r name[24]` | `r plot.id(70)` | `r plot.id(71)` | `r plot.id(72)` | `r s.test(24, "s", res)` | `r s.test(24, "r", res)`|
`r name[25]` | `r plot.id(73)` | `r plot.id(74)` | `r plot.id(75)` | `r s.test(25, "s", res)` | `r s.test(25, "r", res)`|


## Empirical results (all measurements)


```{r test_empirical, echo = FALSE}
## Testing the differences for each distributions
bhatt.coeff.safe <- function(x, y, tol = 1e-16, ...) {
  if(all(x < tol) | all(y < tol)) {
    bhatt.coeff(1, 0)
  } else {
    bhatt.coeff(x, y, ...)
  }
} 
disparity_test <- lapply(all_metrics_empirical_results, lapply, test.dispRity, test = bhatt.coeff.safe)
```

```{r fable_results_empirical_all_metrics, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Plotting parameters
plot.param <- list(cex = 2,
                   col = c("#F7B27E", "#BFE4E3"),
                   border = c("#F65205", "#3E9CBA"),
                   na.cex = 3,
                   scaler = 3)

## Looping through each mini plot for every metric
data_names <- c("Beck and Lee 2014", "Wright 2017", "Marcy et al. 2016",
                "Hopkins et al. 2016", "Jones et al. 2015", "Healy et al. 2019")

for(dataset in 1:length(data_names)){
    for(metric in 1:length(all_metrics)){
        ## Plotting the results
        generate.fable.empirical(data = all_metrics_empirical_results[[dataset]][[metric]],
                                 test = disparity_test[[dataset]][[metric]],
                                 precision = 1e-5, plot.param, dataset = dataset)
    }
}
```

```{r change_plotid_chain, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[3]] <- substitute(chain <- "fable_results_empirical_all_metrics")
```

```{r, echo = FALSE, eval = FALSE}
print.fable(length(all_metrics), byrow = FALSE, ncol = 6, test = FALSE)
```

Measurement      | Beck and Lee 2014  | Wright 2017 | Marcy et al. 2016 | Hopkins et al. 2016 | Jones et al. 2015 | Healy et al. 2019  |
:-----------|----------------|-----------------|------------------|----------------|------------------|----------------|
`r name[1]` | `r plot.id(1)` | `r plot.id(26)` | `r plot.id(51)` | `r plot.id(76)` | `r plot.id(101)` | `r plot.id(126)`|
`r name[2]` | `r plot.id(2)` | `r plot.id(27)` | `r plot.id(52)` | `r plot.id(77)` | `r plot.id(102)` | `r plot.id(127)`|
`r name[3]` | `r plot.id(3)` | `r plot.id(28)` | `r plot.id(53)` | `r plot.id(78)` | `r plot.id(103)` | `r plot.id(128)`|
`r name[4]` | `r plot.id(4)` | `r plot.id(29)` | `r plot.id(54)` | `r plot.id(79)` | `r plot.id(104)` | `r plot.id(129)`|
`r name[5]` | `r plot.id(5)` | `r plot.id(30)` | `r plot.id(55)` | `r plot.id(80)` | `r plot.id(105)` | `r plot.id(130)`|
`r name[6]` | `r plot.id(6)` | `r plot.id(31)` | `r plot.id(56)` | `r plot.id(81)` | `r plot.id(106)` | `r plot.id(131)`|
`r name[7]` | `r plot.id(7)` | `r plot.id(32)` | `r plot.id(57)` | `r plot.id(82)` | `r plot.id(107)` | `r plot.id(132)`|
`r name[8]` | `r plot.id(8)` | `r plot.id(33)` | `r plot.id(58)` | `r plot.id(83)` | `r plot.id(108)` | `r plot.id(133)`|
`r name[9]` | `r plot.id(9)` | `r plot.id(34)` | `r plot.id(59)` | `r plot.id(84)` | `r plot.id(109)` | `r plot.id(134)`|
`r name[10]` | `r plot.id(10)` | `r plot.id(35)` | `r plot.id(60)` | `r plot.id(85)` | `r plot.id(110)` | `r plot.id(135)`|
`r name[11]` | `r plot.id(11)` | `r plot.id(36)` | `r plot.id(61)` | `r plot.id(86)` | `r plot.id(111)` | `r plot.id(136)`|
`r name[12]` | `r plot.id(12)` | `r plot.id(37)` | `r plot.id(62)` | `r plot.id(87)` | `r plot.id(112)` | `r plot.id(137)`|
`r name[13]` | `r plot.id(13)` | `r plot.id(38)` | `r plot.id(63)` | `r plot.id(88)` | `r plot.id(113)` | `r plot.id(138)`|
`r name[14]` | `r plot.id(14)` | `r plot.id(39)` | `r plot.id(64)` | `r plot.id(89)` | `r plot.id(114)` | `r plot.id(139)`|
`r name[15]` | `r plot.id(15)` | `r plot.id(40)` | `r plot.id(65)` | `r plot.id(90)` | `r plot.id(115)` | `r plot.id(140)`|
`r name[16]` | `r plot.id(16)` | `r plot.id(41)` | `r plot.id(66)` | `r plot.id(91)` | `r plot.id(116)` | `r plot.id(141)`|
`r name[17]` | `r plot.id(17)` | `r plot.id(42)` | `r plot.id(67)` | `r plot.id(92)` | `r plot.id(117)` | `r plot.id(142)`|
`r name[18]` | `r plot.id(18)` | `r plot.id(43)` | `r plot.id(68)` | `r plot.id(93)` | `r plot.id(118)` | `r plot.id(143)`|
`r name[19]` | `r plot.id(19)` | `r plot.id(44)` | `r plot.id(69)` | `r plot.id(94)` | `r plot.id(119)` | `r plot.id(144)`|
`r name[20]` | `r plot.id(20)` | `r plot.id(45)` | `r plot.id(70)` | `r plot.id(95)` | `r plot.id(120)` | `r plot.id(145)`|
`r name[21]` | `r plot.id(21)` | `r plot.id(46)` | `r plot.id(71)` | `r plot.id(96)` | `r plot.id(121)` | `r plot.id(146)`|
`r name[22]` | `r plot.id(22)` | `r plot.id(47)` | `r plot.id(72)` | `r plot.id(97)` | `r plot.id(122)` | `r plot.id(147)`|
`r name[23]` | `r plot.id(23)` | `r plot.id(48)` | `r plot.id(73)` | `r plot.id(98)` | `r plot.id(123)` | `r plot.id(148)`|
`r name[24]` | `r plot.id(24)` | `r plot.id(49)` | `r plot.id(74)` | `r plot.id(99)` | `r plot.id(124)` | `#r plot.id(149)`|
`r name[25]` | `r plot.id(25)` | `r plot.id(50)` | `r plot.id(75)` | `r plot.id(100)` | `r plot.id(125)` | `#r plot.id(150)`|