---
title: "Shifting spaces: how to summarise multidimensional spaces occupancy? "
author: "Thomas Guillerme, Mark Puttick, Ariel Marcy, Vera Weisbecker"
bibliography: references.bib
date: "`r Sys.Date()`"
output:
  html_document:
  fig_width: 8
  fig_height: 8
  keep_tex: true
  self_contained: false
  line_numbers: true
  fig_caption: true
  lineno: linenumbers
---

<!-- When  toggling to html, toggle snippet `compilation_html` below to eval = TRUE -->

```{r header, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
## Repeatability note:
## This whole paper is entirely reproducible and compiles as a single document. The data, figures
## and tables are all generated through the code snippets in this file. Note that the first time
## that this paper is compiled, it will generate the data which will take substantial time
## as the shiftingspace_supplementary.Rmd file completion takes ~30 minutes. Subsequent compilations
## will be much faster!

## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerme/moms"); library(moms)

## Create data directory
if(!dir.exists("../data/processed")) {
    dir.create(path="../data/processed", showWarnings = FALSE)
}

## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues <- seq(15, 375, length = n + 1)
    grDevices::hcl(h <- hues, l = 65, c = 100)[1:n]
}

## Setting the datapath
DATA_PATH <- "../data/processed/"

## Checking whether the data exists
if(!all(c("remove_05.Rda", "rarefy_dim_05.Rda") %in% list.files(DATA_PATH))) {
    ## Run The supplementary material (simulation)
    rmarkdown::render("shiftingspace_supplementary_simulation.Rmd", "html_document")    
}
# if(!all(c("palaeo.Rda", "gmm.Rda", "ecol.Rda") %in% list.files(DATA_PATH))) {
#     ## Run The supplementary material
#     rmarkdown::render("shiftingspace_supplementary_empirical.Rmd", "html_document")
# }
```


```{r compilation_html, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[2]] <- substitute(type <- ".png")
```

# Abstract

<!-- Eco/Evo option: -->
Trait multidimensional analysis are now a common tool-kit in ecology and evolution.
Such analysis are based on a multidimensional where each dimension represent a trait (or a transformation thereof) thus summarising all observed and potential trait combinations in a multidimensional trait-space (e.g. a morphospace or an ecospace).
To quantify the overall occupation of such trait-spaces, researchers apply one or more metrics that summarise the way in which organisms inhabit that 'trait-space'.
Using these metrics, researchers have investigated at how trait space occupancy changes through time, in relation to other groups of organisms or in response to global environmental changes (e.g. global warming events or mass extinctions).

Although few well defined space occupancy metrics are used in ecology and evolution, we lack a general description of the majority of those metrics, especially in macroevolution (often called disparity metrics).
Therefore the mathematical and biological meaning of such trait-space occupancy metrics can be vague.
This as consequences when interpreting changes in trait-space occupancy and linking them to biological phenomenon.
For example, does a reduction in volume in trait-space corresponds to episodes of extinction or changes in traits distribution?
Does an overlap in trait-space means overlapping niches?

<!-- Evo only option:
Analysis of the diversity of morphological shapes (aka disparity analysis) are now a common toolkit in macroevolution.
Disparity analysis are based on the analysis of a morphospace which is a multidimensional space summarising all real or potential shapes.
To quantify the overall occupation of morphospace, researchers apply one or more disparity metrics that summarise the way in which taxa inhabit 'trait space'.
Using these metrics of disparity, workers have investigated how taxa morphospace occupancy changes through time, in relation to other clades, or in response to environmental changes, such as a mass extinction.

However, we lack a general description of the majority of disparity metrics, and so the mathematical and biological meaning of disparity metrics is vague.
Thus when we interpret changes in disparity metric values we can only interpret a difference as "a change in multidimensional space occupancy". 

Seperate measures of spatial occupancy can be of particular interest in different studies.
For example, some questions might focus in changes in the volume of the morphospace whilst some other might focus on the position of a group in the morphospace or its density (or all three together).
 -->
<!-- > VW: It seems to me that the review is more important than the tool. 
> VW: So I'd be tempted to go " Here we use X method to review the properties of several commonly used disparity metrics, and introduce two new ones. 
> VW: We do this in the context of a new tool which allows easy replication of the behaviours, and also allows a fully user-controlled interface for allowing the development and testing of metrics on a diversity of datasets"
> VW: Just delete what's hyperbole :-)
> MNP - I agree with these proposed changes.
 -->
Here we propose a user friendly tool to Measure Occupancy in Multidimensional Spaces (moms) that allows to visualise numerical changes of many disparity metrics following space transformations (e.g. contraction, displacement, etc.) as well as visualising a 2D projection of the associated space changes.
This tool thus allows researchers to explore and visualise the properties of disparity metrics under various parameters.
Furthermore, we propose a thorough review of the behaviour of seven common and two new disparity metrics regarding various spacial properties (i.e. distribution, transformation, dimensionality).
As expected, we found that different metrics are more appropriated to be used in different disparity analysis.
<!-- > MNP. Rather than just observing differences, may it be more important to (i) show what different metrics measure and can we actually relate this to underlying biology, and (ii) generate null changes with parameters of interest for a researcher's clade. -->

# Introduction

Life on earth is essentially multidimensional.
Both organisms and their environments are a combination of multiple of traits whether they are phenotypic, physiological, phylogenetic, behavioural, etc.
In fact since the advent of Biology, it has been acknowledged that groups of species and environments share specific, easily recognisable, correlated characteristics (e.g. guilds or biomes).
In fact, a consensus is emerging that organisms or environments should be studied as a set of traits rather than some specific traits in isolation [e.g. @donohue2013, @hopkins2017].
This has led to and explosion of multiple traits studies approaches facilitated by increased computational power and statistical methods [SomeBigDataCite] as well as by the implementation of statistical methods in R packages [@oksanen2007vegan, @adams2013geomorph, @blonder2018, @disprity].

One popular way to analyse all these traits in concert is to use ordination (or dimensionality reduction) techniques [see @legendre2012 for a summary] to create a trait-space of the studied system.
Such trait-space approach has been successfully used across many sub-disciplines in biology to either explore properties of the data or test hypothesis.
For example, in palaeobiology, such trait spaces can be used to look how groups of species change in terms of shared characteristics through time [e.g. @wright2017];
in ecology, it can be used to compare two populations and see whether their overlap can be related to ecological processes such as competition [e.g. @jones2015].
However, different fields use a different set of terms and approaches for such trait-space approaches.
In macroevolution, for example, such trait-spaces are commonly referred to as "morphospaces" (with the whole procedure often being often called "disparity analysis") [e.g. @adams2013geomorph, @marcy2016, @wright2017], while in ecology, such trait-spaces can be called "functional spaces" (and "dissimilarity analysis") [e.g. @diaz2016, @mammola2019].
Nonetheless, they are the same mathematical objects: matrices with columns representing a trait (or a transformation thereof) and rows representing observations (taxon, field site, etc.) [@disprity].

Furthermore, researchers working on ecology or evolution are also using trait-spaces for similar reasons.
Across many sub-disciplines they will look at how their observations occupy the trait-space.
In fact, they often are interested in the same fundamental questions: are two groups overlapping in the trait-space? Are some regions of the trait space not occupied? Or how does specific factors influence the occupancy of the trait-space?
This is achieved in various ways such as using disparity metrics (or indices) in macroevolution [@wills2001, @hopkins2017, @disprity] or comparing hypervolumes in ecology [@donohue2013, @diaz2016, @blonder2018, @mammola2019].
Although these trait-space occupancy metrics are common in ecology and evolution, surprisingly little work has been published on their behaviour [but see @ciampaglio2001, @mammola2019].
This can seem trivial but (unsurprisingly) different space occupancy metrics will capture different aspects of the trait-space.
In many studies in ecology and evolution authors have focused on measuring the volume of the trait-space (or a function thereof; e.g. ellipsoid volume @donohue2013, hypervolume @diaz2016, product of variance @wright2017, Procrustes variance @marcy2016), yet this only represent a single aspects of multidimensional occupancy that exclude other aspects that can be of interest (e.g. the density, @geiger2008; the position @wills2001, @ciampaglio2001; or dissimilarity between groups @oksanen2007vegan, @close2015).
Therefore, this lack of diversity in multidimensional occupancy metrics may restrain the potential of multidimensional analysis.

Here we propose a broad classification of space occupancy metrics as used across ecology and evolution and analyse their power to detect changes in multidimensional space occupancy in empirical and simulated data.
We propose a review of each broad type of space occupancy metrics and a unified terminology aiming to foster easier communication between ecology and evolution.
Unsurprisingly, we found that there is no one-size-fits-all metric and that they are often dependent on the characteristics of the space and the question at hand.
Because they can potentially be as many metrics than they are trait-spaces and studies, it would be a daunting task to propose clear generalities to space occupancy metrics behaviour, therefore, we propose a user friendly tool allowing researchers to design, experiment and visual their own space occupancy metric tailored for their specific project.

## Space occupancy metrics

As mentioned before, we will call trait-spaces any matrix where rows are observations and columns are traits.
This traits can widely vary in number and types.
For example traits could be discrete observations on an organism (e.g. presence of absence of a bone; @beck2014, @wright2017), continuous ones (e.g. leaf area; @diaz2016) or more sophisticated continuous ones (e.g. landmark position; @marcy2016).
Traits can also be relative observations such as community composition (e.g. @jones2015) or even some distance from other observations (e.g. @close2015).
However, regardless of the methodology used to build a trait-space, three main broad occupancy metrics can be measured to investigate how observations are distributed in the trait-space: the volume which will approximate the *amount* of space occupied, the density which will approximate the *distribution* in space and the position which will approximate the *location* in space (Figure 1).
Of course any combination of these three aspects is always possible.


```{r fig_metrics_types, echo = FALSE, fig.height = 3, fig.width = 9, results = 'hide', fig.cap = paste("Figure 1: different type of information captured by disparity metrics. A - Volume (e.g. sum of ranges); B - Density (e.g. average squared pairwise distances); C - Position (e.g. median distance from centroid).") }

set.seed(11)
## Plot space function (utility shortcut)
## The elements
elements <- 10

## Trait space
trait_space <- space.maker(elements, 2, distribution = rnorm)

## Graphical parameters
op <- par(mfrow = c(1,3), bty = "n")

## The volume
plot(trait_space, pch = defaults$pch, main = "A - Volume", xlab = "Trait 1", ylab = "Trait 2")
## The range lines
lines(x = range(trait_space[,1]), y = rep(mean(trait_space[,2]), 2), col = defaults$col1)
lines(x = rep(mean(trait_space[,1]), 2), y = range(trait_space[,2]), col = defaults$col1)

## The density
plot(trait_space, pch = defaults$pch, main = "B - Density", xlab = "Trait 1", ylab = "")
## Plotting the pairwise lines
pair.line <- function(line, trait_space, defaults) {
    lines(x = trait_space[line,1], y = trait_space[line,2], col = defaults$col1)
}
apply(combn(1:elements, 2), 2, pair.line, trait_space, defaults)
points(trait_space, pch = defaults$pch)

## The position
plot(trait_space, pch = defaults$pch, main = "C - Position", xlab = "Trait 1", ylab = "")
## Plotting the center of the space
points(0,0, pch = 13, cex = 2)
arrows(x0 = 0, y0 = 0, x1 = trait_space[,1], y1 = trait_space[,2], code = 0, col = defaults$col1)
points(trait_space, pch = defaults$pch)

par(op)
```

#### 1. The volume

Volume metrics measures the spread of a group in the trait-space.
Such metrics can be interpreted as the amount of the trait-space that is occupied by observations.
Typically, larger values for such metrics indicate more more extreme trait combinations.
For example if group A has a bigger volume than group B, the observations in group A achieve more extreme trait combinations than in group B.
This type of metrics is really common in both ecology (the hypervolume @blonder2018) or in evolution (the sum or product of ranges or variances @wills2001, @ciampaglio2001 or the Procrustes variance @adams2013geomorph).
In general, such metrics are a variation of the hypercube volume.

Although this type of metric is a really good indicator for comparing group's trait-space occupancy, it is only limited to comparing the range of extreme trait-combinations between groups.
Typically, such metrics do not take into account the spread of the observations within a group.
In other words, such metric can make it difficult to determine whether all the observations are on the edge of the volume or whether the volume is simply driven by a couple of extreme observations.
This weakness has been acknowledged in palaeoecology for example for the product of ranges metric (Table 2 - the exact approximation of the hypercube) which is highly sensitive to outliers [@Butler2012] or in ecology with the use of convex-hypervolumes.
However, this can be easily alleviated by using lower confidence intervals (e.g. by only at 95% of the data; @donohue2013 or by using the variance instead of the range; @ciampaglio2001).

#### 2. The density

Density metrics measure the distribution of a group in the trait-space.
Such metrics can be interpreted as an indication of the "packedness" of a group in trait-space.
For example, if group A is more dense than group B, observations in group A tend to be more similar between each other relatively than within group B.
Although this aspect of trait-space occupancy is less routinely measured, it is still often used in both evolution [e.g. the average pairwise distance @geiger2008] or in ecology [e.g. the minimum spanning tree length @oksanen2007vegan].

This type of metrics could be used more often in concert of a volume to allow answering finer hypethises.
For example, if group A has a greater volume than group B but both have the same density, similar mechanisms could be driving both groups trait-space occupancy (e.g. a Brownian motion) but maybe group A is older and had more time to achieve more extreme trait combinations.

#### 3. The position

Position metrics measure where a group lies in trait-space.
Such metrics can be interpreted as an indication of where a group lies in the trait-space either relative to the space itself or relative to another group.
For example, if group A has a different position than group B, observation in group A will have a different trait-combination than group B.

This type of metrics is probably harder to interpret in multidimensional spaces (i.e. beyond left/right and above/below) and hence less often used.
However, when thinking about unidimensional data, this metric is pretty obvious: for example, two groups A or B could have the same variance (i.e. "volume" or spread) with the same number of observations (i.e. the same density) but could have a totally different mean and thus be in different positions.
Although these metrics are more rare in macroevolution (to our knowledge), they have been used in ecology to compare the position of two groups relatively to each other [@mammola2019].


## One metric to rule them all?

In an ideal word, the choice of the metric should be based on the properties of changes in the trait-space of interest while taking into account the properties of the space (number of elements or/and dimensions and distribution of the data on each axis).
However, it is highly unlikely that a such one-size-fits-all metric exist; rather, it is important to assess the properties of a metric relative to the specific data used.

If the question of interest is, say, to look at how trait-space occupancy changes in response to mass extinction, using a single metric to describe one aspect of occupancy (usually the volume) can miss an important part of the picture.
For example, in @halliday2015, the authors demonstrate a change in mammalian trait-space volume as a result from the Cretaceous-Palaeogene extinction (66 million years ago - Mya - Fig. 5 in @halliday2015) yet this one seems to be also accompanied by an apparent shift in the trait-space position due to the extinction of basal mammals (Fig. 8 in @halliday2015).
Such examples can be found throughout the literature in both evolution and ecology on different aspects of trait-space occupancy [e.g. @close2015, @diaz2016].
Figure 2 illustrates this problem where the three aspects (volume, density and position) can be decouple and change in concert or independently.
For this example, both trait space (2D), the observations (8) and the metrics have been optimised for illustration purpose.
In empirical data, the relations can be more complex to discern.
Yet, this illustrates how using only one specific space occupancy metric (e.g. the volume) will fail to capture changes in the trait-spaces C, D, and I.


<!-- In Slater et al 2017: variation in mean squared pairwise distance (density) with two peaks (25 Mya and 5 Mya) corresponding to a less densily packed trait-space (Fig. 5) which might be not optimal to explain the shift in body mass (position) and the increase in variance (volume) for the one trait analysed (body length).
 -->


```{r fig_metric_captures, echo = FALSE, fig.height = 4, fig.width = 8, results = 'hide', fig.cap = paste("Figure 2: Illustration how three different volume (Vol. - Product of ranges), density (Den. - Average nearest neighbour distance) and position (Pos. - Average displacement) metrics capture different aspects of trait-space occupancy.") }

## Making the volume/density/position space
vdp_space <- moms::vdp.make()

## Measuring disparity
vdp_disp <- moms::vdp.dispRity(vdp_space,
                         volume = c(prod, ranges),
                         density = c(mean, neighbours),
                         position = c(mean, displacements))

## Checking the table of volume changes
silent <- moms::vdp.check.table(vdp_disp, vdp_space)

## Plotting the results
moms::vdp.plot(vdp_space, disparity = vdp_disp,
        plot.names = c("A - Base (no changes)",
                       "B - Change in volume",
                       "C - Change in density",
                       "D - Change in position",
                       "E - Change in vol. & den.",
                       "F - Change in vol. & pos.",
                       "I - Change in den. & pos.",
                       "J - All changes"))
```

To alleviate this problem we suggest using multiple metrics to describe trait-space occupancy the same way that multiple statistics are required to describe distributions.
This has also to advantage of allowing finer scale and more precise evolutionary or ecological questions.
For example, during a radiation following a mass extinction event, one could both predict an increase of the volume of the trait-space [e.g. an increase in morphological diversity in birds; @crouch2019] but also a shift if position in the trait-space [e.g. a shift from non-avian to avian dinosaur traits; @crouch2019].
Similarly, in ecology, differences in density of the trait-space while the volume remains constant could indicate some kind of selection pressures on the studied traits (i.e. due to the "narrowness" of trait combination for example in desert plants).

In this paper, we analyse the effect of changes in simulated and empirical trait-spaces on five@@@ commonly used space occupancy metrics as well as proposing two@@@ novel metrics to explicitly measure trait-space density and position.
As expected, we found no one-size-fits-all metric but provide guidance on which metric could be used in which specific scenario.
We also propose a new user friendly software with an intuitive graphical user interface (GUI) to look at the effect of changes in morphospace occupancy on user defined disparity metrics: [`moms`]().


# Methods

We tested the effect of changes in trait-spaces on space occupancy (see [Shifting space](##shifting-spaces)) as a function of spaces dimensions and distributions on the seven@@@ occupancy metrics.
To so, we preformed these following steps (explained in more details below):

1. We simulated seven different spaces with different set of parameters
2. We transformed these spaces by removing 50% of the observations (elements) following four different scenarios: randomly, by limit, by displacement and by density
3. We measured occupancy on the resulted transformed spaces using seven@@@ different metrics
4. We applied the same space occupancy metrics to six@@@ empirical datasets from the literature (covering a range of disciplines and a range of dataset properties).

## Generating spaces

We generated nine spaces using the following parameters:

| space name | size  | distribution(s)                | dimensions variance | correlation |
|------------|-------|--------------------------------|---------------------|-------------|
| Uniform2   |200*2  | Uniform (min = -0.5, max = 0.5)| Constant            | none        |
| Uniform50  |200*50 | Uniform                        | Constant            | none        |
| Uniform150 |200*150| Uniform                        | Constant            | none        |
| Uniform50c |200*50 | Uniform                        | Constant            | Random (between 0.1 and 0.9) |
| Normal2    |200*2  | Normal  (mean = 0, sd = 1)     | Constant            | none        |
| Normal50   |200*50 | Normal                         | Constant            | none        |
| Normal150  |200*150| Normal                         | Constant            | none        |
| Normal50c  |200*50 | Normal                         | Constant            | Random (between 0.1 and 0.9) |
| Random     |200*50 | Normal, Uniform, Lognormal (meanlog = 0, sdlog = 1)| Constant            | none        |
| PCA-like   |200*50 | Normal                         | Lognormal           | none        |
| PCO-like   |200*50 | Normal                         | Normal              | none        |
Table 1: different simulated space distributions.

The different in space sizes (200 $\times$ 2, 200 $\times$ 50 or 200 $\times$ 150) reflects the range of trait-space dimensions in literature: "low-dimension" spaces ($<10$) are common in ecology [@mammola2019] whereas high dimension spaces ($>100$) are common in macroevolution [@hopkins2017].
We used a different range of distributions (uniform, normal or random) to test the effect of observation distributions on the metrics.
It can be assumed that many trait-spaces are normally distributed [e.g. @diaz2016, @cooney2017, @healy2019] albeit with some varying levels of correlation [e.g. @jones2015, @wright2017] reflected in the correlation parameter.
For simplification, in the seven first spaces, we used a constant variance per dimensions (i.e. all dimensions have a similar variance) but we used a decreasing variance for the "PCA-like" (Principal Component Analysis) and "PCO-like" (Principal Coordinates Analysis) spaces to mimic the ranking of variance in ordination technique.
Empirically variance tends to decrease log-normally in PCAs [e.g. @marcy2016, healy2019] and more normally in PCOs [e.g.  @close2015, @wright2017].


## Shifting spaces {#shifting-spaces}

We reduced the spaces by removing 50% of elements each time using the following algorithms

* **Randomly:** by randomly removing 50% of elements (Fig. 3A).

* **Limit:** by removing all elements with a distance from the centre (mean point) of the space lower or greater than a radius $\rho$ (where \$rho$ is selected such that 50% elements are selected) (Fig. 3B).

* **Displacement:** by removing points similarly as above but using the distance from the furthest point from the centre (max point - Fig. 3C).

* **Density:** by removing any pairs of point with a distance $D$ from each other where (where $D$ is selected such that 50% elements are selected - Fig. 3D).

The algorithm to select $\rho$ or $D$ is described in more detail in in the [Appendix](#Appendix_algorithm_reduce).
We used mirror results for the limit, displacement and density reduction (i.e. the opposite data was also used).
For example, when increasing the **density** by removing the 50% most distant elements, we also run a transformation by removing 50% of the *closests* elements.
This resulted in seven different transformation by reducing the space: randomly, by limit (positive or negative), by displacement (positive or negative) or by density (increased or decreased).


```{r fig_reduce_space, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 2: different type of space reduction. Each panel displays 50% of the data points removed (in grey) using the following algorithm: A - randomly; B - by limit (removing data from the centre); C - by displacement (removing data from one side of the plot); and D - by density (removing data to increase or decrease the density).")}
set.seed(42)
op <- par(mfrow = c(2,2), bty = "n")
## The elements
elements <- 300
## The amount to remove
remove <- 0.5

## Trait space
trait_space <- space.maker(elements, 2, distribution = rnorm)
# ## The original space
# plot(trait_space, pch = 20, xlim = c(-3, 3), ylim = c(-3, 3),
#      main = "A - Original space", xlab = "Trait 1", ylab = "Trait 2")

## The randomly removed space
random_rm <- reduce.space(trait_space, type = "random", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, random_rm, main = "A - Random removal", defaults)

## The limit removal
limit_rm <- reduce.space(trait_space, type = "limit", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, limit_rm, main = "B - Limit", defaults)

## The displacement removal
displace_rm <- reduce.space(trait_space, type = "displacement", remove = 0.5)
## Moving the space to the upper right corner (positive)
plot.space(trait_space, displace_rm, main = "C - Displacement", defaults)

## The density removal
density_rm <- reduce.space(trait_space, type = "density", remove = 0.5)
## Plotting the reduction
plot.space(trait_space, density_rm, main = "D - Density", defaults)

## List of removals (for plotting the tables later)
typical_removal <- list(space = trait_space,
remove = list("random" = random_rm,
              "limit(min)" = limit_rm,
              "limit(max)" = !limit_rm,
              "displace(positive)" = displace_rm,
              "displace(negative)" = !displace_rm,
              "density(high)" = density_rm,
              "density(low)" = !density_rm))
```

We repeated each of the data generation process (generating the space and reducing it) 20 times to obtain variance on the simulated parameters.

## Disparity metrics

We then measured seven different metrics on the resulting transformed spaces:

Name          | Definition                      | Captures | Origin           | `dispRity` function
:-------------|:-------------------------------:|----------|------------------|------|
Average squared distance | $\left(\sqrt{\sum_{i}^{n}{({q}_{i}-p_{i})^2}}\right)^{2}\times \frac{1}{n}$ | Density and Volume | @geiger2008 | `mean(dispRity::pairwise.dist(matrix)^2)`
Procrustes variance | $\sum_{i}^{m=dn}{M_{m}^2} \times \frac{1}{d}$ | Density and Volume | @adams2013geomorph | `sum(matrix^2)/nrow(matrix)`
Ellipsoid volume | $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ | Volume | @donohue2013 | `dispRity::ellipse.volume(matrix)`
Product of ranges | $\prod_{i}^{d}{\|max(k_i) - min(k_i)\|}$ | Volume | @wills2001 | `prod(dispRity::ranges(matrix))`
Sum of ranges | $\sum_{i}^{d}{\|max(k_i) - min(k_i)\|}$ | Volume | @wills2001 | `sum(dispRity::ranges(matrix))`
Product of variances | $\prod_{i}^{d}{\sigma^{2}{k_i}}$ | Volume | @wills2001 | `prod(dispRity::variances(matrix))`
Sum of variances | $\sum_{i}^{d}{\sigma^{2}{k_i}}$ | Volume | @wills2001 | `sum(dispRity::variances(matrix))`
Average displacement | $\frac{\sqrt{\sum_{i}^{n}{({k}_{n})^2}}}{\sqrt{\sum_{i}^{n}{({k}_{n}-Centroid_{k})^2}}}$ | Position | This paper | `mean(dispRity::displacements(matrix))`
Average nearest neighbour distance |$min\left(\sqrt{\sum_{i}^{n}{({q}_{i}-p_{i})^2}}\right)\times \frac{1}{n}$ | Density | This paper | `mean(dispRity::neighbours(matrix))`








```{r, all_metric, echo = FALSE}
## All metrics (exploration)
metrics_list <- list("av.pairwise" = function(matrix) return(mean(pairwise.dist(matrix)^2)),
                    "Procrustes" = function(matrix) return(sum(matrix^2)/nrow(matrix)),
                    "Volume" = ellipse.volume,
                    "PoR" = c(prod, ranges),
                    "SoR" = c(sum, ranges),
                    "PoV" = c(prod, variances),
                    "SoV" = c(sum, variances),
                    # "PoQ" = c(prod, quantiles),
                    # "SoQ" = c(sum, quantiles),
                    "av.displac" = c(mean, displacements),
                    "av.neighbo" = c(mean, neighbours))

metric_names <- c("Average squared\ndistance", "Procrustes\nvariance", "Ellipsoid\nvolume", 
                  "Product of\nRange", "Sum of\nRanges", "Product of\nVariances",
                  "Sum of\nVariances", "Mean\ndisplacement", "Mean nearest\nneighbours distance")
## Shorter vector name version (for clarity in the fable)
name <- gsub("\n", " ", metric_names)
#"Product of Quantiles", "Sum of Quantiles", 
```

```{r, complex_metrics_example, echo = FALSE, eval = FALSE}
complex_metric <- list("med.mana.dist.centre" = list(c(median, centroids),
                       centroid = 0, method = "manhattan"))
```

## Effect of dimensionality

> VW: I understand teh below but only just and it is hard to read. I can help simplify.

Furthermore, high dimensional spaces ($>> 3$) are subject to the "curse of multidimensionality" [@cursedimensionality].
This is link to the fact that the more dimensions, the more sparse the data becomes.
In other words the probability of two points A and B overlapping in *n* dimensions is the probability of point A having the same value as point point B on dimension 1, 2, 3, etc...
Thus this probability decreases as a product of the number of dimensions.
This "curse" can make the interpretation of high dimensional data counter-intuitive.

For example, the product of range is expected to decrease with the number of dimensions, regardless of the number of elements on each dimensions.
This is counter-intuitive because:
1) first, the range of values on a dimension is bounded to the number of elements on this dimensions (i.e. increasing the number of observations on one dimension can only increase the range, never decrease it);
2) second, products are a classical example of the "curse" where multiplying a higher number of dimensions with a non-null probability of having a range of zero will irremediably result in 0 (i.e. an infinite number of positive values by 0 is still 0).


```{r fig_curse, echo = FALSE, fig.height = 8, fig.width = 8, results = 'hide', fig.cap = paste("Figure 3: Effect of the curse of dimensionality on different scaled disparity metrics. Metrics with containing products (ellipsoid volume, product of ranges or variances) are classically affected by the \"curse\" as they decrease more or less rapidly with an increase in dimensions. Note especially how the Product of Range actually increases and then decreases with the number of dimensions.")}
set.seed(42)
## Make a 50 dimensions space
space <- space.maker(300, 50, rnorm, scree = rev(cumsum(rep(1/50, 50))))

## Function wrappers
disparity <- function(metric, dimensions, space) {
    fast.disparity(group = 1:nrow(space), space = space, metric = metric, rare.dim = dimensions)
}
lapply.fun <- function(dimensions, metrics_list, space) {
    unlist(lapply(metrics_list, disparity, dimensions = dimensions, space = space))
}

## Curse results
curse_results <- sapply(2:50, lapply.fun, metrics_list = metrics_list, space = space)
curse_results <- apply(curse_results, 1, function(x) return(x/max(x)))

## Empty plot
hue_colors <- default.palette(ncol(curse_results))
par(bty = "n")
plot(NULL, xlim = c(2,50), ylim = c(0,1),
xlab = "Dimensions (decreasing variance)",
ylab = "Scaled disparity")
for(metric in 1:ncol(curse_results)) {
    lines(curse_results[, metric], col = hue_colors[metric])
}
legend("bottomleft", metric_names, lty = 1, col = hue_colors)
```

## Measuring the effect of space reduction, distribution and dimensionality on the metric

### Scaling disparity

To measure the effect of space reduction, distribution and dimensionality on the metric, we first scaled the metric to be relative to the non reduced space for each dimension distributions or number of dimensions.
To do so, we subtracted the observed disparity with no space reduction (i.e. base disparity) to all the disparity measurements of the reduced spaces and then divided it by the resulting maximum observed disparity.
This way, our disparity metrics where scaled between -1 and 1 with a value of 0 indicating no effect of the space reduction and $>0$ and $<0$ respectively indicating an increase or decrease in the disparity metric value.

### Measuring the effect of the metric

TODO: probably simply do a Bhattacharyya coefficient to see how much the disparity from the reduced space overlaps with the randomly reduced disparity.
High values will indicate low effect and low values, high effects.
TODO: probably flip that metric over so that low values = low effects and high values = high effects

$\text{Effect of metric} = 1 - BC(metric_{focal}, metric_{random})$

### Measuring the effect of space and dimensionality

To measure the effect of the different space distributions we run two analysis of variance: each individual disparity metric function of the different seven different spaces and each individual disparity metric function of the 10 different rarefaction levels.
The results (_p_-values) are reported in Table 3.

TODO: report the full tests

> VW:Here or perhaps in the results, I think there needs to be an explicit expectation of what "good" behaviour is for each metric. Something like an expectation - for the disparity metric to work, what should it do? E.g. If you apply the "limit" scenario, the squared distance measure is expected to go down for the black dots and up for the grey dots. 


## Empirical examples

We analysed the effect of the different space occupancy metrics on six different studies from palaeontology, geometric morphometrics, and ecology to cover a broad representative view of fields using trait-spaces analysis.

<!-- Here, we design a new method for assessing the usefulness of various measures of trait space occupation to fields as diverse as palaeontology, ecology, and phylogenetics. To demonstrate the applicability of our disparity assessments to a diversity of data modalities, we apply our approach to examples of phylogenetic character matrices, multivariate ecological data, and procrustes shapes (geometric morphometrics) -->

We used the dataset of each study 


Study | Field | Taxonomic Group | Type of data | Trait-space | Dimensions | groups size | type of question |
------|-------|-----------------|--------------|-------------|-------------|------------------|
@beck2014 | palaeontology | Mammalia | discrete phylogenetic data | Ordination of a distance matrix (PCO) | @@@ | 52 crown vs. 50 stem | Are crown mammals more disparate than stem mammals?|
@wright2017 | palaeontology | Crinoidea | discrete phylogenetic data | Ordination of a distance matrix (PCO) | @@@ | 16 before vs. 23 after | Is there a difference in disparity before and after the Ordovician mass extinction?|
@marcy2016 | geometrics morphometrics | Rodentia | 2D landmarks | Ordination of a Procrustes Superimposition (PCA) | @@@ |@@@ vs. @@@ | @@@Are these two groups of Gophers different?|
@@@ | geometrics morphometrics | @@@ | 3D landmarks | Ordination of a Procrustes Superimposition (PCA) | @@@ |@@@ vs. @@@ | @@@|
@jones2015 | Ecology | Plantae | | | | | |
@@@ | Ecology | @@@ | Continuous data | Ordination (PCA) | @@@ | @@@ vs. @@@ | @@@|


We re-analised the data from the original studies in three different ways to obtain equivalent trait-spaces with two groups of roughly the same size.

* **Palaeontology** [@beck2014, @wright2017]: 1) calculating a distance matrix (using MORD distances) and 2) ordinating the distance matrix using a PCoA (or PCO).
* **Geometric morphometrics** [@marcy2016]: 1) running a General Procrustes Analysis and 2) ordinating the Procrustes data using a PCA.
* **Ecology** [@jones2015]: 1) scaling the data and 2) ordinating the scaled data using a PCA.

Each procedure is described in more details in the supplementary materials.
<!-- TODO: vignette to generate the trait-spaces -->


# Results

```{r loading_results, echo = FALSE}
## Loading the results
remove_05 <- load.results("remove_05")
rarefy_dim_05 <- load.results("rarefy_dim_05")
```

```{r running_tests, echo = FALSE}
## Anova function
anova.fun <- function(data) {return(aov(glm(disparity ~ factor, data = data)))}

## Running the tests
test_space05 <- test.metrics(remove_05, test = anova.fun, scale = TRUE)
test_raref05 <- test.metrics(rarefy_dim_05, test = anova.fun, scale = TRUE)

#TODO: Generate the aov tables in the supplementary

```


```{r fable_results, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Making a list of parameters for each mini plot
plot.param <- list(scaler = 0.5,
                   bg.col = "black",
                   col = c("grey", "orange", "blue"),
                   quantiles = c(95, 50),
                   cent.tend = median,
                   pch = 19)

## Looping through each mini plot for every metric
for(metric in 1:length(metrics_list)) {
    generate.fable.plot(data = remove_05, metric = metric, what = "limits", plot.param = plot.param)
    generate.fable.plot(data = remove_05, metric = metric, what = "displa", plot.param = plot.param)
    generate.fable.plot(data = remove_05, metric = metric, what = "densit", plot.param = plot.param)
}

```


Table 3: Results of the effect of space reduction, space dimensions distributions and dimensions number of the different disparity metrics.
Figures: the dots represent the median disparity values, the solid and dashed line respectively the 50% and 95% confidence intervals. Results in grey are the random 50% reduction, in orange the negative reductions (i.e. discarding the black dots in Figure 2) and in blue the positive reductions (i.e. discarding the grey dots in Figure 2). _p_-values for distribution effect and dimensions effect represents respectively the effect of the ANOVAs disparity ~ distributions and disparity ~ dimensions (Signif. codes:  0 '\*\*\*' 0.001 '\*\*' 0.01 '\*' 0.05 '.' 0.1 ' ' 1).

Metric      | Volume change  | Position change | Density change | Distribution effect | Dimensions effect |
:-----------|----------------|-----------------|----------------|---------------------|-------------------|
`r name[1]` |`r plot.id(1)`  |`r plot.id(2)`   |`r plot.id(3)`  | `r s.test(1, "s")`  | `r s.test(1, "r")`|
`r name[2]` |`r plot.id(4)`  |`r plot.id(5)`   |`r plot.id(6)`  | `r s.test(2, "s")`  | `r s.test(2, "r")`|
`r name[3]` |`r plot.id(7)`  |`r plot.id(8)`   |`r plot.id(9)`  | `r s.test(3, "s")`  | `r s.test(3, "r")`|
`r name[4]` |`r plot.id(10)` |`r plot.id(11)`  |`r plot.id(12)` | `r s.test(4, "s")`  | `r s.test(4, "r")`|
`r name[5]` |`r plot.id(13)` |`r plot.id(14)`  |`r plot.id(15)` | `r s.test(5, "s")`  | `r s.test(5, "r")`|
`r name[6]` |`r plot.id(16)` |`r plot.id(17)`  |`r plot.id(18)` | `r s.test(6, "s")`  | `r s.test(6, "r")`|
`r name[7]` |`r plot.id(19)` |`r plot.id(20)`  |`r plot.id(21)` | `r s.test(7, "s")`  | `r s.test(7, "r")`|
`r name[8]` |`r plot.id(22)` |`r plot.id(23)`  |`r plot.id(24)` | `r s.test(8, "s")`  | `r s.test(8, "r")`|
`r name[9]` |`r plot.id(25)` |`r plot.id(26)`  |`r plot.id(27)` | `r s.test(9, "s")`  | `r s.test(9, "r")`|

Table caption goes here


```{r running_empirical, echo = FALSE}
## Loading the data
palaeo <- load.results("palaeo")
gmm <- load.results("gmm")
ecol <- load.results("ecol")

## List of spaces
trait_spaces <- unlist(list(palaeo, gmm, ecol), recursive = FALSE)

## Measuring space occupancy for all spaces and all metrics
apply.disparity <- function(space, metrics) {
    lapply(metrics, function(metric, space) dispRity(boot.matrix(space), metric), space)
}
disparities <- lapply(trait_spaces, apply.disparity, metrics_list)

# ## Wrapper for the bhatt.coeff function to return NA below a precision number
# ## (to avoid comparing effectively 0 vs 0 distributions)
# bhatt.coeff.safe <- function(x, y, bw = bw.nrd0, precision, ...) {
#     if(all(x < precision) || all(y < precision)) {
#         return(NA)
#     } else {
#         return(bhatt.coeff(x, y, bw = bw, ...))
#     }
# }

# ## Testing the probability of overlap
# disparity_bc <- lapply(disparities, lapply, test.dispRity,
#                            test = bhatt.coeff.safe, precision = 1e-50)

## Testing the differences for each distributions
disparity_t_test <- lapply(disparities, lapply, test.dispRity, test = t.test)
```

```{r fable_results_empirical, fig.show='hide', echo=FALSE, fig.height=3, fig.width=3}
## Plotting parameters
plot.param <- list(cex = 1,
                   col = c("#F7B27E", "#BFE4E3"),
                   border = c("#F65205", "#3E9CBA"),
                   na.cex = 3)

## Looping through each mini plot for every metric
data_names <- c("Beck and Lee 2014", "Wright 2017", "Marcy et al. 2016",
                "Somebody", "Jones et al. 2015", "Somebody")

for(dataset in 1:length(data_names)){
    for(metric in 1:length(metric_names)){
        ## Plotting the results
        generate.fable.empirical(disparities[[dataset]][[metric]], disparity_t_test[[dataset]][[metric]],
                                 precision = 1e-5, plot.param)
    }
}
```

```{r change_plotid_chain, echo = FALSE, eval = TRUE}
## Changing defaults
body(plot.id)[[3]] <- substitute(chain <- "fable_results_empirical")
```


Metric      | Beck and Lee 2014  | Wright 2017 | Marcy et al. 2016 | Somebody       | Jones et al. 2015 | Somebody       |
:-----------|----------------|-----------------|------------------|----------------|------------------|----------------|
`r name[1]` |`r plot.id(1)`  |`r plot.id(10)`  |`r plot.id(19)`   |`r plot.id(28)` |`r plot.id(37)`   |`r plot.id(46)` |
`r name[2]` |`r plot.id(2)`  |`r plot.id(11)`  |`r plot.id(20)`   |`r plot.id(29)` |`r plot.id(38)`   |`r plot.id(47)` |
`r name[3]` |`r plot.id(3)`  |`r plot.id(12)`  |`r plot.id(21)`   |`r plot.id(30)` |`r plot.id(39)`   |`r plot.id(48)` |
`r name[4]` |`r plot.id(4)`  |`r plot.id(13)`  |`r plot.id(22)`   |`r plot.id(31)` |`r plot.id(40)`   |`r plot.id(49)` |
`r name[5]` |`r plot.id(5)`  |`r plot.id(14)`  |`r plot.id(23)`   |`r plot.id(32)` |`r plot.id(41)`   |`r plot.id(50)` |
`r name[6]` |`r plot.id(6)`  |`r plot.id(15)`  |`r plot.id(24)`   |`r plot.id(33)` |`r plot.id(42)`   |`r plot.id(51)` |
`r name[7]` |`r plot.id(7)`  |`r plot.id(16)`  |`r plot.id(25)`   |`r plot.id(34)` |`r plot.id(43)`   |`r plot.id(52)` |
`r name[8]` |`r plot.id(8)`  |`r plot.id(17)`  |`r plot.id(26)`   |`r plot.id(35)` |`r plot.id(44)`   |`r plot.id(53)` |
`r name[9]` |`r plot.id(9)`  |`r plot.id(18)`  |`r plot.id(27)`   |`r plot.id(36)` |`r plot.id(45)`   |`r plot.id(54)` |

Some table caption goes here










# Discussion


* Average squared distances: all right metric for measuring density (increase/decrease picked up). Not great for volume (i.e. volume decreases when the morphospace becomes hollow). Not at all good for position (both negative/positive are picked up the same way). Affected by the distributions but not the dimensions.

* Procrustes variances: similar to the average squared distances although even worth for position (often close to random changes). Clearly not affect by the distributions and the dimensions.

* Ellispsoid volume: Not great for picking up anything, probably because of the curse of multidimensionality. This metric is clearly affected both by the distributions and the number of dimensions. This metric is probably best used with a low number of dimensions ($<10$).

* Product of ranges: Same as the ellipsoid volume.

* Sum of ranges: Picks a only negative change in volume (but not hollowness) and bat for measuring changes in position and density. This metric is clearly affected by the distributions and also by the number of dimensions (albeit less clearly).

* Product of variances: Positive decrease in volume (hollowness) is picked up. But not much signal from change in position or density. This metric is clearly affected by the distributions and the dimensions.

* Sum of variances: Similar behaviour than the Procrustes variance. Picks up great differences in volume. Bad for position on average, not bad for density. 

* Mean displacement: Picks up reduction in volume (increased metric), not great for change in position and density. This metric is clearly affected by both the distributions and the number of dimensions.

* Mean nearest neighbour distances: Picks up reduction in volume. Not great for change in position. OK for change in density. This metric is affected by the distributions but not the number of dimensions.

## Caveat of the analysis

* Data not simulated using evolutionary models (TODO: maybe TODO?)
* Response to reduction is often more specific to distribution/dimensions/reduction for each metric and these fine differences are not displayed here.
*

# Conclusion

There are non one-size-fits-all disparity metric.
It is important to use the disparity metric that will best capture the problem at hand (e.g. if interested in the position of the morphospace, don't use the product of variances).

Furthermore, all these different metrics are "disparity metrics" even though they represent widely different aspects.
We thus suggest that future disparity analysis authors abstain from using the term "disparity" to designate the "disparity metric" but rather use "position in the morphospace", or, directly the name of the used metric to avoid confusion.


# Acknowledgements

Thanks to Natalie Jones for helping finding/sorting the ecological data.
Money from the ARC.
ARC DP170103227, FT180100634



# References





# Supplementary material

## Algorithm for selecting the parameters to reduce the space ($radius$, $displacement$, $density$) {#Appendix_algorithm_reduce}

We used a recursive algorithm for selecting the parameter that removes $P$ elements: 80%, 50% or 20%.

1. Select a random reduction parameters $R$.
2. Remove elements from the space using $R$ resulting in $P'$ removed elements.
If the remaining number of elements is to the required proportion $P$ ; exit the algorithm;
Else go to 3.
3. Get the different between the proportion of removed elements $P'$ and $P$.
If the difference is positive set the increment parameter $R$ to $R = 1.1 \times R$, then go to 2.
Else set $R = 0.9 \times R$, then go to 2.

The algorithm is implemented in the `optimise.parameter` function in [reduce.space_fun.R](@@@).
