---
title: "Shifting spaces: how to summarise multidimensional spaces occupancy? "
author: "Thomas Guillerme, Mark Puttick, Ariel Marcy, Vera Weisbecker"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Supplementary material 2: empirical data

```{r, echo = FALSE}
## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerme/moms"); library(moms)
## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues = seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[1:n]
}
```

# Empirical data

## Creating the trait spaces

The following sections contains the methodology and the code to generate the six empirical trait-spaces analysed in the paper.
The protocol for creating each space is detailed below.

### Palaeontological (discrete) data

For the discrete morphological data, we calculated a distance matrix from the original discrete morphological matrices from @beck2014 and @wright2017 using the maximum observable distance [@lloyd2016, @lehmann2019] that we then ordinated using a classical multidimensional scaling (MDS; [@torgerson1965, gower1966, cailliez1983]) by adding a constant **c\*** to make the distances Euclidean ($d_{MORD} + c* = d_{Euclidean}; @cailliez1983).
This resulted in ***k*** eigenvectors for each matrix where ***k*** is equal to $n-1$ taxa. 

```{r}
## Creating the trait-spaces
beck_traitspace <- Claddis.ordination("../data/matrices/Beck2014.nex", distance = "MORD")
wright_traitspace <- Claddis.ordination("../data/matrices/Wright2017.nex", distance = "MORD")
```

For both datasets we created two groups, the crown versus the stem mammals for @beck2014 and the crinoids occuring before and after the Ordovician-Silurian boundary (455-430 million years ago - Mya).

```{r}
## Loading the tree from Beck and Lee 2014
beck_tree <- read.nexus("../data/groups/Beck2014.tre")

## Selecting the crown and stem mammals
groups_beck <- crown.stem(beck_tree)

## Loading the first and last occurence data from Wright 2017
wright_ages <- read.csv("../data/groups/Wright2017.csv")

## Separating taxa that went extinct half way through the Ordovician-Silurian extinction (442.5 Mya)
groups_wright <- ifelse(wright_ages$LAD < 422.5, TRUE, FALSE)
groups_wright <- list("Before" = as.character(wright_ages$species[which(groups_wright)]),
                      "After" =  as.character(wright_ages$species[which(!groups_wright)]))

## Creating the dispRity objects
beck_traitspace <- custom.subsets(beck_traitspace, group = groups_beck)
wright_traitspace <- custom.subsets(wright_traitspace, group = groups_wright)

## Saving the data for the main analysis
palaeo <- list("beck" = beck_traitspace, "wright" = wright_traitspace)
save.results(palaeo)
```


### Geometric morphometric (landmark) data

For the geometric morphometric data (landmark datasets) we used...

```{r, eval = FALSE}
## Using the geomorph package
if(!require(geomorph)) install.packages("geomorph"); library(geomorph)

## Loading the landmark data

## Performing a Procrustes superimposition
 <- geomorph::gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)

## Ordinating the data
 <- geomorph.ordination()

```

```{r, eval = FALSE}
## Making the groups

## Saving the data for the main analysis
gmm <- list("beck" = beck_traitspace, "wright" = wright_traitspace)
save.results(gmm)
```

### Ecological (continuous) data

```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
```

```{r, eval = FALSE}
## Making the groups

## Saving the data for the main analysis
ecol <- list("beck" = beck_traitspace, "wright" = wright_traitspace)
save.results(ecol)
```



# References