---
title: "Shifting spaces: how to summarise multidimensional spaces occupancy? "
author: "Thomas Guillerme, Mark Puttick, Ariel Marcy, Vera Weisbecker"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Supplementary material 2: empirical data

```{r header_empirical, echo = FALSE}
## Loading the packages
if(!require(devtools)) install.packages("devtools")
if(!require(knitr)) install.packages("knitr"); library(knitr)
if(!require(rmarkdown)) install.packages("rmarkdown"); library(rmarkdown)
if(!require(ape)) install.packages("ape"); library(ape)
if(!require(dispRity)) install.packages("dispRity"); library(dispRity)
if(!require(geomorph)) install.packages("geomorph"); library(geomorph)
if(packageVersion("dispRity") < "1.2.4") {
    ## dispRity must be above v1.2.3
    devtools::install_github("TGuillerme/dispRity"); library(dispRity)
}
if(!require(moms)) devtools::install_github("TGuillerme/moms"); library(moms)
## Setting the default parameters for the space plots
defaults <- list(pch = 20,
                 xlim = c(-3, 3),
                 ylim = c(-3, 3),
                 col1 = "grey",
                 col2 = "black",
                 xlab = "Trait",
                 ylab = "Trait",
                 cex = 1)
## Generating the default palette
default.palette <- function(n) {
    hues = seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[1:n]
}
```

# Empirical data

## Creating the trait spaces

The following sections contains the methodology and the code to generate the six empirical trait-spaces analysed in the paper.
The protocol for creating each space is detailed below.

### Palaeontological (discrete) data

For the discrete morphological data, we calculated a distance matrix from the original discrete morphological matrices from @beck2014 and @wright2017 using the maximum observable distance [@lloyd2016, @lehmann2019] that we then ordinated using a classical multidimensional scaling (MDS; [@torgerson1965, gower1966, cailliez1983]) by adding a constant **c\*** to make the distances Euclidean ($d_{MORD} + c* = d_{Euclidean}; @cailliez1983).
This resulted in ***k*** eigenvectors for each matrix where ***k*** is equal to $n-1$ taxa. 

```{r load_claddis_empirical}
## Creating the trait-spaces
beck_traitspace <- Claddis.ordination("../data/raw/Beck2014.nex", distance = "MORD")
wright_traitspace <- Claddis.ordination("../data/raw/Wright2017.nex", distance = "MORD")
```

For both datasets we created two groups, the crown versus the stem mammals for @beck2014 and the crinoids occuring before and after the Ordovician-Silurian boundary (455-430 million years ago - Mya).

```{r group_claddis_empirical}
## Loading the tree from Beck and Lee 2014
beck_tree <- read.nexus("../data/raw/Beck2014.tre")

## Selecting the crown and stem mammals
groups_beck <- crown.stem(beck_tree)

## Adding the four stem mammals not in the tree
missing_stem <- rownames(beck_traitspace)[which(is.na(match(rownames(beck_traitspace), 
                                                            beck_tree$tip.label)))]
groups_beck$stem <- c(groups_beck$stem, missing_stem)

## Loading the first and last occurence data from Wright 2017
wright_ages <- read.csv("../data/raw/Wright2017.csv")

## Separating taxa that went extinct half way through the Ordovician-Silurian extinction (442.5 Mya)
groups_wright <- ifelse(wright_ages$LAD < 422.5, TRUE, FALSE)
groups_wright <- list(#"all" = as.character(wright_ages$species),
                      "before" = as.character(wright_ages$species[which(groups_wright)]),
                      "after" =  as.character(wright_ages$species[which(!groups_wright)]))

## Creating the dispRity objects
beck_traitspace <- custom.subsets(beck_traitspace, group = groups_beck)
wright_traitspace <- custom.subsets(wright_traitspace, group = groups_wright)

## Saving the data for the main analysis
palaeo <- list("beck" = beck_traitspace, "wright" = wright_traitspace)
save.results(palaeo)
```


### Geometric morphometric (landmark) data

For the continuous geometric morphometric data, we performed a generalised Procrustes analysis [@adams2013geomorph], using 2D [@marcy2016] or 3D [@hopkins2016] landmark coordinates sets (with semi-landmarks, i.e. sliders).
We then ordinated the resulted Procrustes superimposed data using a principal component analysis [PCA; @hotelling1933] and divided the data into two groups for each dataset (see below and in the main text).

#### Marcy et al. 2016

This dataset consists 2D landmarks for 552 specimens [25 fixed landmarks and 42 semi-landmark curves; @marcy2016].
The dataset is then divided into two genera[@marcy2016].

```{r load_marcy_empirical}
## Read in TPS coordinates
marcy_data <- readland.tps("../data/raw/Marcy2016.TPS", specID = "imageID", warnmsg = FALSE)

## Remove duplicated points for ventral crania semi-landmarks (artefact of the landmarking process)
marcy_data <- marcy_data[-c(1, 2, 26, 29, 30, 41, 42, 45, 46, 51, 58, 60, 61, 68, 69, 78, 79, 85), , ]

## Read in semilandmark definitions from Marcy et al 2016
sliders <- read.csv("../data/raw/Marcy2016.csv", header = TRUE, colClasses = "integer")

## Get the specimens names
marcy_names <- strsplit(names(marcy_data[1, 1, ]), "_")

## Find non-unique specimens (they have a suffix, i.e. 7 elements)
duplicated <- ifelse(unlist(lapply(marcy_names, length)) == 7, TRUE, FALSE)

## Remove duplicated specimens
marcy_data <- marcy_data[, , !duplicated]
marcy_names <- marcy_names[!duplicated]

## Procrustes superimposition
marcy_gpa <- gpagen(marcy_data, Proj = TRUE, ProcD = TRUE, curves = sliders, print.progress = FALSE)
```

#### Hopkins et al. 2016 (ontogenic sequences)

This dataset consists of 63 3D landmarks for 46 specimens [from 23 individuals, 3 of which have estimated landmarks, @hopkins2016].
The data is then separated into juveniles and adults [based on the specimens centroids sizes; @hopkins2016].

```{r load_hopkins_empirical, warning = FALSE}
## Read in TPS coordinates
hopkins_data <- readland.tps("../data/raw/Hopkins2016.tps", specID = "ID", warnmsg = FALSE)

## Remove duplicated points (repeated coordinates reported in Appendix 3 of original paper)
hopkins_data <- hopkins_data[-c(63, 67, 55, 65), , ] 

## Estimate missing landmarks using TPS method (used in orignial paper)
hopkins_data <- estimate.missing(hopkins_data, "TPS")

## Procrustes superimposition
hopkins_gpa <- gpagen(hopkins_data, print.progress = FALSE)
```

##### Groups

We divided the dataset from @marcy2016 into two groups corresponding to the two genera: *Thomomys* and *Megascapheus*; and the dataset from @hopkins2016 into two groups, adult and juvenile based on the log centroid size threshold of 2.3 [i.e. juveniles have a log centroid size < 2.3; @hopkins].

```{r group_gmm_empirical}
## Selecting the generas (first element in the name list)
marcy_groups <- unlist(lapply(marcy_names, function(x) x[[1]]))

## Renaming the generas
marcy_groups <- as.factor(ifelse(marcy_groups == "T", "Thomomys", "Megascapheus"))

## Creating the geomorph data.frames
marcy_data <- geomorph.data.frame(marcy_gpa, marcy_groups)

## Setting the threshold for distinguising adults/juveniles in the hopkins data
size_threshold <- 2.3

## Creating the groupins
hopkins_groups <- ifelse(log(hopkins_gpa$Csize) > size_threshold, "adult", "juvenile")

## Creating the geomorph data.frames
hopkins_data <- geomorph.data.frame(hopkins_gpa, hopkins_groups)
```

```{r morphospace_gmm_empirical, eval = TRUE, warning = FALSE}
## Creating the trait space
marcy_traitspace <- dispRity::geomorph.ordination(marcy_data)
hopkins_traitspace <- dispRity::geomorph.ordination(hopkins_data)

## Saving the data for the main analysis
gmm <- list("marcy" = marcy_traitspace, "hopkins" = hopkins_traitspace)
save.results(gmm)
```

### Ecological data

#### Jones et al. 2015 data (species composition data)

For the @jones2015 dataset, we converted the spreadsheet "aspen & grass richness & abund" from the data from @jones2015 available at @jones2015data into a .csv file (now available in `/data/raw`).
We then selected only the `wholeplot` entries and separated the studied plot between `aspen` and `grassland` and converted the species presence data into a presence/absence matrix.
From this matrix we calculated a Jaccard distance matrix [@oksanen2007vegan] that we then ordinated using a classical multidimensional scaling (MDS; [@torgerson1965, cailliez1983]) by adding a constant **c\*** to make the distances Euclidean ($d_{MORD} + c* = d_{Euclidean}; @cailliez1983).
This resulted in ***k*** eigenvectors for each matrix where ***k*** is equal to $n-1$ field sites. 

```{r load_jones_empirical, eval = TRUE}
## Reading the community composition data from Jones et al 2015
jones_communities <- read.csv("../data/raw/Jones2015.csv", stringsAsFactors = FALSE)

## Selecting only the whole plots and the plot number, habitat and species
jones_whole_com <- jones_communities[which(jones_communities$mptype == "wholeplot"), c(1,2,3)]

## Separate each plot into aspen/grassland
jones_whole_com$plot <- paste(jones_whole_com[,1], jones_whole_com$habitat, sep = ".")

## Making the community matrix
plot_names <- unique(jones_whole_com$plot)
sp_codes <- unique(jones_whole_com$species.code)
community <- matrix(NA, nrow = length(plot_names),
                        ncol = length(sp_codes),
                        dimnames = list(plot_names, sp_codes))

## Populate the matrix
for(one_plot in rownames(community)) {
    ## Get the community composition
    composition <- match(colnames(community),
                        jones_whole_com[which(jones_whole_com$plot == one_plot), 3])
    ## Fill the plot row with presence/absence data
    community[which(rownames(community) == one_plot), ] <- ifelse(is.na(composition), 0, 1)
}

## Calculate the distance matrix
distance_matrix <- as.matrix(vegan::vegdist(community, method = "jaccard"))

## Ordinating the data
jones_ordination <- stats::cmdscale(distance_matrix,
                                    k = ncol(distance_matrix)- 1, add = TRUE)$points
```

#### Healy et al. 2019 data (life history traits data)

For the @healy2019 dataset, we used a processed version of the data available on the [GitHub repository associated to the paper](https://github.com/healyke/Healy_et_al_2019_Animal_Life_History).

The procedure from @healy2019 consists of analysing matrix population models and extracting life history traits (survival rate, age at first reproduction, generation time, reproduction rate, life expectancy post sexual maturity and reproduction spread).
To take into account the effect of body mass and phylogeny, @healy2019 used the residual of the six traits from general linear mixed models [@mcmcglmm; @multree].
See @healy2019 for the exact description of the procedure.

We used the [PCA_main_2019_may_28](https://github.com/healyke/Healy_et_al_2019_Animal_Life_History/blob/master/Main_PCA_analysis/PCA_main_2019_may_28.Rmd) script to obtain the data corrected for body mass and phylogeny with six variables: the survival residuals (`"SD_mort"`), the age at first reproduction (`"La_r"`), the generation time (`"gen_r"`), the mean reproductive rate (`"M_repo"`), the life expectancy post sexual maturity (`"M_suv"`) and the spread of reproduction (`"gini_r"`).

```{r healy_pca}
## Load the processed data
healy_data <- read.csv("../data/raw/Healy2019.csv")

## Cleaning up the data
healy_traits <- healy_data[, -c(1, 8, 9)]

## Ordinating the traits
healy_ordination <- prcomp(healy_traits)$x
```


### Making the ecological trait-space groups

We separated the @jones2015 dataset into `aspen` and `grassland` groups and we separated the @healy2019 dataset into `ectotherms` `endotherms`:

```{r group_ecol_empirical, eval = TRUE, warning = FALSE}
## Making the groups
jones_groups <- list("aspen" = grep("aspen", rownames(jones_ordination)),
                     "grassland" = grep("grassland", rownames(jones_ordination)))
healy_groups <- list("ectotherms" = which(healy_data[, 9] == "ecto"),
                     "endotherms" = which(healy_data[, 9] == "endo"))


## Making the dispRity objects
jones_traitspace <- custom.subsets(jones_ordination, group = jones_groups)
healy_traitspace <- custom.subsets(healy_ordination, group = healy_groups)

## Saving the data for the main analysis
ecol <- list("jones" = jones_traitspace, "healy" = healy_traitspace)
save.results(ecol)
```


### Exporting the datasets for shiny

```{r demo_data_empirical}
## The list of datasets
demo_data <- c(palaeo, gmm, ecol)
save(demo_data, file = "../data/demo_data.rda")
```

# Running the disparity calculations

```{r metrics_list_empirical, echo = FALSE}
## Loading the list of metrics
source("list.of.metrics.R")
```

```{r running_empirical, echo = FALSE}
## Loading the data
trait_spaces <- demo_data

## Measuring space occupancy for all spaces and all metrics
apply.disparity <- function(space, metrics) {
    lapply(metrics, function(metric, space) dispRity(boot.matrix(space, bootstraps = 500), metric),
           space)
}
empirical_results <- lapply(trait_spaces, apply.disparity, metrics_list)

## Results list
save.results(empirical_results)
```


# References